---
title: "F3 — Training vs Test Error"
engine: knitr
---

```{=html}
<script>document.body.classList.add("foundations");</script>
```

```{r}
#| context: setup
#| include: false
library(ggplot2)
library(dplyr)
set.seed(42)
f_true <- function(x) sin(pi * x)
```

```{webr-r}
#| context: setup
#| include: false
library(ggplot2)
library(dplyr)
set.seed(42)
f_true <- function(x) sin(pi * x)
```

# Concept: Why Training Error is Optimistic

## Optimism of training error

When we fit $\hat{f}$ on training data and then evaluate on the **same** data, the estimate is **optimistically biased**. The model has already adapted to the specific noise in the training set.

The expected optimism is:

$$\text{Optimism} = \mathbb{E}[\text{Test MSE}] - \mathbb{E}[\text{Train MSE}] = \frac{2}{n} \sum_{i=1}^n \text{Cov}(\hat{y}_i, y_i)$$

For OLS with $p$ predictors: $\text{Optimism} = 2p\sigma^2 / n$. This is why AIC and Mallow's $C_p$ add a penalty of $2p\hat\sigma^2 / n$ to the training RSS.

## Train/test split as the direct approach

The simplest approach: split data once into training and test sets (e.g., 70/30). Fit on training, evaluate on test.

**Advantage:** Simple, unbiased estimate of test error.

**Disadvantage:** Wastes data; test MSE estimate has high variance (depends on which observations ended up in the test set).

Cross-validation (F5) improves on this by reusing all data.

---

# Quiz

```{=html}
<div class="sta230-quiz" data-question="You fit a linear model with p=10 predictors on n=50 training observations. The training MSE is 4.2. Approximately what is the optimism (expected gap to test MSE)?">
  <div class="q-opt" data-correct="false">0.42 — roughly 10% of the training MSE</div>
  <div class="q-opt" data-correct="false">4.2 — equal to the training MSE itself</div>
  <div class="q-opt" data-correct="true">Cannot be computed — σ² is unknown; the formula is 2pσ²/n</div>
  <div class="q-opt" data-correct="false">Zero — OLS is unbiased so training error matches test error</div>
  <div class="q-exp">The optimism formula for OLS is 2pσ²/n. With σ² unknown, we must estimate it. If σ̂² ≈ 4.2 (MSE ≈ variance), then optimism ≈ 2·10·4.2/50 = 1.68 — a 40% underestimate of true test error. The key insight: more predictors → more optimism → training MSE more misleading.</div>
</div>

<div class="sta230-quiz" data-question="You randomly split your dataset 70/30 and get a test MSE of 12.3. You repeat the split with a different random seed and get 9.1. What is the main problem demonstrated here?">
  <div class="q-opt" data-correct="false">The model is overfit — training error is much lower than test error</div>
  <div class="q-opt" data-correct="false">The training set is too large, causing unstable coefficient estimates</div>
  <div class="q-opt" data-correct="true">High variance in the test MSE estimate from a single random split</div>
  <div class="q-opt" data-correct="false">The true test error is undefined without a fixed target population</div>
  <div class="q-exp">A single train/test split gives an unbiased but noisy estimate of the true test error. The variance of the estimate depends on which observations happened to be in the test set. Cross-validation averages over many splits, reducing this variance substantially.</div>
</div>
```

---

# Worked Example

```{r}
#| label: train-test-optimism
set.seed(42)
n <- 100
x <- sort(runif(n, 0, 1))
y <- f_true(x) + rnorm(n, 0, 0.4)

# Fit models of increasing degrees and compute both train and test error
n_test <- 500
x_te <- seq(0, 1, length.out=n_test)
y_te <- f_true(x_te) + rnorm(n_test, 0, 0.4)

degrees <- 1:12
errors <- lapply(degrees, function(d) {
  fit <- lm(y ~ poly(x, d))
  xp_test <- poly(x_te, d, coefs=attr(poly(x, d), "coefs"))
  data.frame(
    degree = d,
    train  = mean(residuals(fit)^2),
    test   = mean((y_te - cbind(1, xp_test) %*% coef(fit))^2)
  )
})
err_df <- do.call(rbind, errors)

ggplot(err_df, aes(x=degree)) +
  geom_line(aes(y=train), colour="#2980b9", linewidth=1.3) +
  geom_point(aes(y=train), colour="#2980b9", size=2.5) +
  geom_line(aes(y=test),  colour="#e74c3c", linewidth=1.3) +
  geom_point(aes(y=test),  colour="#e74c3c", size=2.5) +
  geom_hline(yintercept=0.16, linetype="dashed", colour="#888") +
  scale_x_continuous(breaks=degrees) +
  labs(title="Training vs test error on a single dataset",
       subtitle="Blue = training  |  Red = test  |  Dashed = σ² = 0.16",
       x="Polynomial degree", y="MSE") +
  theme_minimal()
```

---

# Guided Practice

::: {.webr-loading-note}
**R is loading in the background** — exercises below will be interactive once ready.
:::

```{webr-r}
#| label: guided-split
set.seed(42)
n <- 200
x <- sort(runif(n, 0, 1))
y <- f_true(x) + rnorm(n, 0, 0.4)

# TASK 1: Split 70/30 and compute test MSE for poly(1), poly(3), poly(8)
train_idx <- sample(n, 140)
train_dat <- data.frame(x=x[train_idx], y=y[train_idx])
test_dat  <- data.frame(x=x[-train_idx], y=y[-train_idx])

for (d in c(1, 3, 8)) {
  fit <- lm(y ~ poly(x, d), data=train_dat)
  xp <- poly(test_dat$x, d, coefs=attr(poly(train_dat$x, d), "coefs"))
  test_mse <- mean((test_dat$y - cbind(1, xp) %*% coef(fit))^2)
  cat(sprintf("poly(%d) — Test MSE: %.4f\n", d, test_mse))
}

# TASK 2: Repeat with seed=123 and seed=456 — how much does test MSE vary?
```

::: {.callout-tip collapse="true"}
## Hints
- Change `set.seed(42)` to `set.seed(123)` and `set.seed(456)` and re-run
- Record the test MSE for poly(3) across the three seeds — note the variability
- This variability is exactly what cross-validation reduces
:::

::: {.callout-note}
## Further reading
James et al. (2021). *An Introduction to Statistical Learning*. — Chapter 5.1
:::
