---
title: "F5 — Cross-Validation in Practice"
engine: knitr
---

```{=html}
<script>document.body.classList.add("foundations");</script>
```

```{r}
#| context: setup
#| include: false
library(ggplot2)
library(glmnet)
set.seed(42)
```

```{webr-r}
#| context: setup
#| include: false
library(ggplot2)
library(glmnet)
set.seed(42)
```

# Concept: Using CV for Model Selection

## The one-standard-error rule

When selecting the best model via CV, the **minimum CV error** may not be the best choice. The **1-SE rule** selects the simplest model whose CV error is within one standard error of the minimum.

$$d^* = \min \{d : \text{CV}(d) \leq \text{CV}(d_{\min}) + \text{SE}(d_{\min})\}$$

This guards against selecting an unnecessarily complex model when a simpler one performs nearly as well.

## CV in glmnet: choosing λ

For ridge and LASSO regression, `cv.glmnet()` automates k-fold CV over a grid of $\lambda$ values and returns:

- `lambda.min`: the $\lambda$ minimising CV error
- `lambda.1se`: the largest $\lambda$ within 1 SE of the minimum (more regularisation, simpler model)

In practice, `lambda.1se` is often preferred for parsimony.

## Pitfalls

1. **Information leakage**: never standardise predictors on all data before CV — standardise inside each fold
2. **Nested CV**: if you select both the model class *and* its hyperparameters, you need outer CV for honest evaluation
3. **Classification**: use stratified folds to preserve class proportions across folds

---

# Quiz

```{=html}
<div class="sta230-quiz" data-question="The 1-SE rule selects a simpler model than the one minimising CV error. Why might this be preferred?">
  <div class="q-opt" data-correct="false">It always gives lower test MSE</div>
  <div class="q-opt" data-correct="true">Models within 1 SE of each other are statistically tied — the simpler one is preferred by Occam's razor</div>
  <div class="q-opt" data-correct="false">It corrects for optimism in the CV estimate</div>
  <div class="q-opt" data-correct="false">It is required by glmnet and cannot be disabled</div>
  <div class="q-exp">CV error estimates have uncertainty (measured by the SE across folds). If two models' CV errors are within 1 SE of each other, we cannot distinguish them statistically. The 1-SE rule then picks the simpler one — Occam's razor applied to model selection. This is often more robust when deployed on new data.</div>
</div>

<div class="sta240-quiz" data-question="You run cv.glmnet() for LASSO and see lambda.min = 0.01, lambda.1se = 0.08. What does this tell you?">
  <div class="q-opt" data-correct="false">The minimum CV error occurs at lambda = 0.08</div>
  <div class="q-opt" data-correct="true">lambda.1se = 0.08 is the most penalised model still within 1 SE of the CV minimum — sparser than at 0.01</div>
  <div class="q-opt" data-correct="false">The model at lambda = 0.01 has higher test error than at 0.08</div>
  <div class="q-opt" data-correct="false">lambda.1se is always larger than lambda.min by definition of LASSO</div>
  <div class="q-exp">lambda.min = 0.01 minimises the cross-validated MSE. lambda.1se = 0.08 is the largest lambda whose CV MSE is within 1 SE of the minimum — meaning it produces a sparser (more zeros) but statistically comparable model. Higher lambda → more shrinkage → fewer non-zero coefficients in LASSO.</div>
</div>
```

---

# Worked Example: cv.glmnet on Hitters

```{r}
#| label: cv-lasso-hitters
library(ISLR2)
library(glmnet)

data(Hitters)
Hitters <- na.omit(Hitters)

X <- model.matrix(Salary ~ ., data=Hitters)[, -1]
y <- Hitters$Salary

set.seed(42)
cv_lasso <- cv.glmnet(X, y, alpha=1, nfolds=10)

cat("lambda.min:", round(cv_lasso$lambda.min, 4), "\n")
cat("lambda.1se:", round(cv_lasso$lambda.1se, 4), "\n")
cat("\nNumber of non-zero coefficients:\n")
cat("  at lambda.min:", sum(coef(cv_lasso, s="lambda.min") != 0) - 1, "\n")
cat("  at lambda.1se:", sum(coef(cv_lasso, s="lambda.1se") != 0) - 1, "\n")

plot(cv_lasso)
```

```{r}
#| label: cv-ridge-hitters
cv_ridge <- cv.glmnet(X, y, alpha=0, nfolds=10)
cat("Ridge lambda.min:", round(cv_ridge$lambda.min, 4), "\n")
cat("Ridge lambda.1se:", round(cv_ridge$lambda.1se, 4), "\n")

# Compare LASSO vs Ridge test MSE
cat("\nTest MSE (LASSO, lambda.min):",
    round(min(cv_lasso$cvm), 2), "\n")
cat("Test MSE (Ridge, lambda.min):",
    round(min(cv_ridge$cvm), 2), "\n")
```

---

# Guided Practice

::: {.webr-loading-note}
**R is loading** — exercises interactive once ready.
:::

```{webr-r}
#| label: guided-cv-glmnet
# TASK: Apply cv.glmnet (LASSO) to predict Salary from Hitters
# and answer the questions below
library(glmnet)

# Simulated dataset with same structure (ISLR2 not available in WebR by default)
set.seed(42)
n <- 263; p <- 19
X_sim <- matrix(rnorm(n*p), n, p)
colnames(X_sim) <- paste0("X", 1:p)
# True model: only X1, X3, X7 matter
beta_true <- c(50, 0, 30, rep(0, 5), -20, rep(0, 10))
y_sim <- 500 + X_sim %*% beta_true + rnorm(n, 0, 80)

# TASK 1: Run 10-fold CV LASSO
cv_fit <- cv.glmnet(X_sim, y_sim, alpha=1, nfolds=10)
plot(cv_fit)

# TASK 2: How many non-zero coefficients at lambda.min vs lambda.1se?
cat("Non-zero at lambda.min:", sum(coef(cv_fit, s="lambda.min") != 0) - 1, "\n")
cat("Non-zero at lambda.1se:", sum(coef(cv_fit, s="lambda.1se") != 0) - 1, "\n")

# TASK 3: Which predictors are selected at lambda.1se?
coef_1se <- coef(cv_fit, s="lambda.1se")
selected <- rownames(coef_1se)[coef_1se[,1] != 0]
cat("Selected at lambda.1se:", paste(selected[-1], collapse=", "), "\n")
# Do they match the true model (X1, X3, X7)?
```

::: {.callout-tip collapse="true"}
## Hints
- `coef(cv_fit, s="lambda.1se")` returns a sparse matrix — use `!=0` to find non-zero entries
- The true model has non-zero coefficients at X1 (β=50), X3 (β=30), X9 (β=−20)
- Does LASSO recover them? Compare with lambda.min
:::

::: {.callout-note}
## Further reading
James et al. (2021). *An Introduction to Statistical Learning*. — Chapter 6.2 (LASSO + CV) and Chapter 5.1 (CV theory)
:::
