---
title: "C1 — Logistic Regression"
engine: knitr
---

```{=html}
<script>document.body.classList.add("classification");</script>
```

```{r}
#| context: setup
#| include: false
library(ggplot2)
library(dplyr)
swp <- readRDS(here::here("data/swp.rds"))
```

```{webr-r}
#| context: setup
#| include: false
library(ggplot2)
library(dplyr)
swp <- readRDS("https://johlobs.github.io/STA230-App/data/swp.rds")
```

# Concept: Classification and the Logistic Model

## From regression to classification

In classification the outcome $Y$ is categorical. For binary $Y \in \{0, 1\}$, we model the **probability** of belonging to class 1:

$$p(\mathbf{x}) = P(Y = 1 \mid \mathbf{X} = \mathbf{x})$$

We then classify: $\hat{Y} = 1$ if $\hat{p}(\mathbf{x}) > 0.5$ (or another threshold).

## The logistic model

Logistic regression models $p(\mathbf{x})$ via the logit link:

$$\log\frac{p(\mathbf{x})}{1 - p(\mathbf{x})} = \beta_0 + \beta_1 x_1 + \cdots + \beta_p x_p$$

Equivalently:

$$p(\mathbf{x}) = \frac{e^{\mathbf{x}^\top\boldsymbol{\beta}}}{1 + e^{\mathbf{x}^\top\boldsymbol{\beta}}}$$

The coefficients $e^{\beta_j}$ are **odds ratios** — the multiplicative change in odds of $Y=1$ per unit increase in $x_j$.

## Class imbalance

When one class is rare (e.g., only 11% have $Y=0$ in SWP_CTW), a naive classifier that always predicts the majority class achieves high accuracy but is useless. Alternatives:

- Use **sensitivity** and **specificity** instead of accuracy
- Report **AUC** (area under the ROC curve)
- Adjust the decision threshold away from 0.5
- Resample: oversample the minority class (SMOTE) or undersample the majority

---

# Quiz

```{=html}
<div class="sta230-quiz" data-question="Logistic regression models P(Y=1|X). Why not just use linear regression and threshold at 0.5?">
  <div class="q-opt" data-correct="false">Linear regression cannot handle binary outcomes mathematically</div>
  <div class="q-opt" data-correct="true">Linear regression can predict outside [0,1] and assumes a linear effect on probability, not on the log-odds</div>
  <div class="q-opt" data-correct="false">Logistic regression always gives better predictions than linear regression for binary outcomes</div>
  <div class="q-opt" data-correct="false">Linear regression requires normally distributed errors, which binary outcomes violate in the coefficients</div>
  <div class="q-exp">Linear regression fitted to a binary outcome (the 'linear probability model') can give predicted probabilities <0 or >1 and has heteroskedastic errors. More fundamentally, the effect of a predictor on probability should be non-linear: near p=0 or p=1, a unit change in X should have little effect; near p=0.5 the effect is largest. The logit link captures this non-linearity correctly.</div>
</div>

<div class="sta230-quiz" data-question="In SWP_CTW, 89% of participants have SWP=1. A model that always predicts SWP=1 achieves 89% accuracy. Why is this misleading?">
  <div class="q-opt" data-correct="false">Accuracy of 89% is always considered a good model</div>
  <div class="q-opt" data-correct="true">It never identifies the at-risk 11% — sensitivity for SWP=0 is zero; the minority class is ignored</div>
  <div class="q-opt" data-correct="false">The model is overfit because training and test accuracy are both 89%</div>
  <div class="q-opt" data-correct="false">Accuracy is only valid when classes are perfectly balanced</div>
  <div class="q-exp">With 89% majority class, predicting '1' for everyone achieves 89% accuracy with zero insight. In occupational health, correctly identifying the 11% who do NOT sustain work (SWP=0) is exactly the clinical goal — those are the people needing intervention. Sensitivity (recall for the minority class) and AUC are far more informative than accuracy in imbalanced settings.</div>
</div>

<div class="sta230-quiz" data-question="The logistic coefficient for CTW is β=0.08. What is the odds ratio, and what does it mean?">
  <div class="q-opt" data-correct="false">OR = 0.08 — a one-unit increase in CTW multiplies the odds of SWP=1 by 0.08</div>
  <div class="q-opt" data-correct="true">OR = exp(0.08) ≈ 1.08 — a one-unit increase in CTW multiplies the odds of SWP=1 by 1.08</div>
  <div class="q-opt" data-correct="false">OR = 0.08 + 1 = 1.08 — odds increase by 8 percentage points</div>
  <div class="q-opt" data-correct="false">OR = exp(0.08)² — the OR must be squared for binary outcomes</div>
  <div class="q-exp">The log-odds is linear in X with slope β. A one-unit increase in CTW changes the log-odds by 0.08, which means the odds are multiplied by exp(0.08) ≈ 1.083. Since CTW ranges ~66–118, a 10-unit increase gives OR = exp(0.8) ≈ 2.23 — the odds of SWP=1 more than doubles. Always report OR = exp(β) with a confidence interval [exp(β − 1.96·SE), exp(β + 1.96·SE)].</div>
</div>
```

---

# Worked Example: Predicting Sustainable Work Participation

```{r}
#| label: data-overview
cat("Dataset: SWP_CTW — Sustainable Work Participation\n")
cat("n =", nrow(swp), " | Outcome: SWP (1 = sustains work)\n")
cat("\nClass distribution:\n")
print(table(SWP = swp$SWP))
cat("\nClass balance:", round(mean(swp$SWP == 1) * 100, 1), "% SWP=1\n")
```

```{r}
#| label: logistic-fit
fit <- glm(SWP ~ Age + Sex + Fulltime + CTW + JobDemands + JobControl +
             JobSupport + MentalWB, data = swp, family = binomial)
summary(fit)
```

```{r}
#| label: odds-ratios
#| fig-width: 8
#| fig-height: 4.5
# Odds ratios with 95% CI
or_df <- data.frame(
  variable = names(coef(fit))[-1],
  OR  = exp(coef(fit)[-1]),
  lower = exp(confint(fit)[-1, 1]),
  upper = exp(confint(fit)[-1, 2])
)

ggplot(or_df, aes(x = reorder(variable, OR), y = OR)) +
  geom_point(size = 3, colour = "#2980b9") +
  geom_errorbar(aes(ymin = lower, ymax = upper), width = 0.2, colour = "#2980b9") +
  geom_hline(yintercept = 1, linetype = "dashed", colour = "#888") +
  coord_flip() +
  labs(title = "Odds ratios for SWP=1 (logistic regression)",
       subtitle = "OR > 1: increases odds of SWP=1  |  OR < 1: decreases odds",
       x = NULL, y = "Odds Ratio (95% CI)") +
  theme_minimal()
```

```{r}
#| label: confusion-matrix
# Default threshold = 0.5
pred_prob <- predict(fit, type = "response")
pred_class <- ifelse(pred_prob > 0.5, 1, 0)

cm <- table(Predicted = pred_class, Actual = swp$SWP)
print(cm)
cat("\nAccuracy:   ", round(mean(pred_class == swp$SWP) * 100, 1), "%\n")
cat("Sensitivity:", round(cm[2,2] / sum(cm[,2]) * 100, 1), "% (correctly identified SWP=1)\n")
cat("Specificity:", round(cm[1,1] / sum(cm[,1]) * 100, 1), "% (correctly identified SWP=0)\n")
```

---

# Guided Practice

::: {.webr-loading-note}
**R is loading in the background** — exercises below will be interactive once ready.
:::

```{webr-r}
#| label: guided-explore
# TASK 1: Check class imbalance and explore predictors
cat("Class distribution:\n")
print(table(SWP = swp$SWP))
cat("\nMean CTW by SWP:\n")
print(tapply(swp$CTW, swp$SWP, mean))
cat("\nMean MentalWB by SWP:\n")
print(tapply(swp$MentalWB, swp$SWP, mean))
```

```{webr-r}
#| label: guided-logistic
# TASK 2: Fit logistic regression and inspect coefficients
fit <- glm(SWP ~ Age + Sex + Fulltime + CTW + JobDemands +
             JobControl + JobSupport + MentalWB,
           data = swp, family = binomial)

# Odds ratios
cat("Odds ratios:\n")
print(round(exp(coef(fit)), 4))

# TASK 3: Which predictors are most strongly associated with SWP?
# (Compare z-values / p-values)
cat("\nCoefficient summary:\n")
print(round(summary(fit)$coefficients, 4))
```

```{webr-r}
#| label: guided-threshold
# TASK 4: Try a lower threshold for the minority class (SWP=0)
# Default threshold = 0.5 rarely catches the minority class well
pred_prob <- predict(fit, type = "response")

for (thresh in c(0.80, 0.85, 0.90, 0.95)) {
  pred <- ifelse(pred_prob > thresh, 1, 0)
  cm   <- table(Predicted = pred, Actual = swp$SWP)
  sens <- if (nrow(cm) == 2) cm[1,1] / sum(cm[,1]) else 0
  spec <- if (nrow(cm) == 2) cm[2,2] / sum(cm[,2]) else 1
  cat(sprintf("Threshold=%.2f  Acc=%.1f%%  Sens(SWP=0)=%.1f%%  Spec(SWP=1)=%.1f%%\n",
              thresh,
              mean(pred == swp$SWP) * 100,
              sens * 100, spec * 100))
}
```

::: {.callout-tip collapse="true"}
## What to look for
- At threshold=0.5, sensitivity for SWP=0 is likely very low — why?
- As you lower the threshold, sensitivity for SWP=0 increases but specificity decreases
- This trade-off is exactly what the ROC curve (C6) visualises
:::

---

# Retrieval Challenge

```{webr-r}
#| label: challenge
# Q1: Train/test split — compare training and test accuracy
set.seed(42)
tr <- sample(nrow(swp), round(0.7 * nrow(swp)))
te <- setdiff(seq_len(nrow(swp)), tr)

fit_tr <- glm(SWP ~ Age + Sex + Fulltime + CTW + JobDemands +
                JobControl + JobSupport + MentalWB,
              data = swp[tr, ], family = binomial)

acc <- function(m, newdat) {
  p <- predict(m, newdata = newdat, type = "response")
  mean(ifelse(p > 0.5, 1, 0) == newdat$SWP)
}
cat("Training accuracy:", round(acc(fit_tr, swp[tr,]) * 100, 1), "%\n")
cat("Test accuracy:    ", round(acc(fit_tr, swp[te,]) * 100, 1), "%\n")

# Q2: Null model accuracy (always predict majority class)
cat("Null accuracy:    ", round(max(mean(swp$SWP==1), mean(swp$SWP==0)) * 100, 1), "%\n")

# Q3: What does this tell you about using accuracy as the metric here?
```

---

# Why-Questions

::: {.callout-caution collapse="true"}
## Q: "Can I use logistic regression with 80 predictors?"

Yes, but with caution. With p=80 and n=10000, OLS-style logistic regression is technically feasible (n >> p), but many of those 80 variables may be redundant questionnaire items. Better to either (a) use penalised logistic regression (LASSO with glmnet, family="binomial"), or (b) first create composite scores (as was done here with CTW, JobDemands etc.) from the raw items. Blindly including all raw items risks overfitting and obscures interpretation.
:::

::: {.callout-note}
## Further reading
James et al. (2021). *An Introduction to Statistical Learning*. — Chapter 4.1–4.3
:::
