---
title: "C2 — LDA & kNN"
engine: knitr
---

```{=html}
<script>document.body.classList.add("classification");</script>
```

```{r}
#| context: setup
#| include: false
library(ggplot2)
library(dplyr)
library(MASS)
library(class)
swp <- readRDS(here::here("data/swp.rds"))
set.seed(42)
tr <- sample(nrow(swp), round(0.7 * nrow(swp)))
train <- swp[tr, ]; test <- swp[-tr, ]
```

```{webr-r}
#| context: setup
#| include: false
library(ggplot2)
library(dplyr)
library(MASS)
library(class)
swp <- readRDS("https://johlobs.github.io/STA230-App/data/swp.rds")
set.seed(42)
tr <- sample(nrow(swp), round(0.7 * nrow(swp)))
train <- swp[tr, ]; test <- swp[-tr, ]
```

# Concept: Generative Classifiers and Distance Methods

## Linear Discriminant Analysis (LDA)

LDA models the class-conditional distributions $P(\mathbf{X} \mid Y=k)$ as multivariate Gaussians with the **same covariance matrix** $\boldsymbol{\Sigma}$. Via Bayes' theorem, this gives a **linear** decision boundary. QDA relaxes equal-covariance, giving a quadratic boundary but estimating more parameters.

## k-Nearest Neighbours (kNN)

kNN is non-parametric — no distributional assumption:

1. For a new point $\mathbf{x}_0$, find the $k$ closest training points
2. Assign the majority class among those neighbours

Small $k$ = flexible, low bias, high variance. Large $k$ = smooth, high bias, low variance. Optimal $k$ found by cross-validation.

| Method | Boundary | Assumption | Scale-sensitive? |
|--------|----------|------------|-----------------|
| Logistic | Linear | None | No |
| LDA | Linear | Gaussian, equal Σ | No |
| QDA | Quadratic | Gaussian, different Σ | No |
| kNN | Non-linear | None | **Yes — must standardise** |

---

# Quiz

```{=html}
<div class="sta230-quiz" data-question="LDA assumes equal covariance matrices across classes. What happens if this is violated?">
  <div class="q-opt" data-correct="false">LDA becomes equivalent to logistic regression</div>
  <div class="q-opt" data-correct="false">LDA fails completely and cannot produce predictions</div>
  <div class="q-opt" data-correct="true">The linear boundary is misspecified — QDA or a flexible method may perform better</div>
  <div class="q-opt" data-correct="false">LDA automatically corrects by estimating separate covariance matrices</div>
  <div class="q-exp">If classes have different covariance structures, the true Bayes boundary is quadratic, not linear. LDA's linear boundary will then be suboptimal. QDA estimates a separate Σ_k per class, giving a quadratic boundary. However, QDA has more parameters and needs more data. With limited data, LDA's lower variance may still outweigh the bias from the wrong boundary shape.</div>
</div>

<div class="sta230-quiz" data-question="You use kNN with k=1 on training data. What is the training error?">
  <div class="q-opt" data-correct="true">Zero — each training point is its own nearest neighbour</div>
  <div class="q-opt" data-correct="false">Equal to the Bayes error rate</div>
  <div class="q-opt" data-correct="false">Equal to 1/n</div>
  <div class="q-opt" data-correct="false">Undefined — kNN cannot be evaluated on training data</div>
  <div class="q-exp">With k=1, every training point is its own nearest neighbour and is classified perfectly — training error is 0. This is extreme overfitting. Test error with k=1 is much higher. As k increases, the model smooths out and test error first falls then rises — the same U-shape as the bias-variance tradeoff.</div>
</div>

<div class="sta230-quiz" data-question="Why must predictors be standardised before applying kNN?">
  <div class="q-opt" data-correct="false">kNN requires all predictors to follow a normal distribution so that Euclidean distance is valid</div>
  <div class="q-opt" data-correct="true">High-range predictors dominate Euclidean distance regardless of their actual predictive importance</div>
  <div class="q-opt" data-correct="false">Standardisation only speeds up computation — the same neighbours are found regardless of scale</div>
  <div class="q-opt" data-correct="false">kNN requires integer-valued predictors because Euclidean distance is undefined for continuous inputs</div>
  <div class="q-exp">kNN uses Euclidean distance. If Age ranges 18–70 and JobControl ranges 85–120, a 1-unit difference in Age contributes more to distance than a 1-unit difference in JobControl — even if JobControl is more informative. Standardising (subtract mean, divide by SD) puts all predictors on equal footing and is essential before kNN.</div>
</div>
```

---

# Worked Example

```{r}
#| label: lda-fit
lda_fit  <- lda(SWP ~ Age + CTW + JobDemands + JobControl + JobSupport + MentalWB,
                data = train)
lda_pred <- predict(lda_fit, newdata = test)
cm_lda   <- table(Predicted = lda_pred$class, Actual = test$SWP)
print(cm_lda)
cat("\nTest accuracy:      ", round(mean(lda_pred$class == test$SWP)*100, 1), "%\n")
cat("Sensitivity (SWP=0):", round(cm_lda[1,1] / sum(cm_lda[,1])*100, 1), "%\n")
cat("Specificity (SWP=1):", round(cm_lda[2,2] / sum(cm_lda[,2])*100, 1), "%\n")
```

```{r}
#| label: knn-k-selection
#| fig-width: 8
#| fig-height: 4
pred_vars <- c("Age","CTW","JobDemands","JobControl","JobSupport","MentalWB")
mu  <- colMeans(train[, pred_vars])
sds <- apply(train[, pred_vars], 2, sd)
X_tr <- scale(train[, pred_vars], center = mu, scale = sds)
X_te <- scale(test[, pred_vars],  center = mu, scale = sds)

# Use a subset for speed during rendering (n=7000 kNN is slow)
set.seed(7)
sub <- sample(nrow(X_tr), 1500)
X_tr_s <- X_tr[sub, ]; y_tr_s <- train$SWP[sub]

ks        <- c(1, 5, 10, 25, 50, 100, 200)
test_err  <- sapply(ks, function(k) mean(knn(X_tr_s, X_te, cl = y_tr_s, k=k) != test$SWP))

ggplot(data.frame(k = ks, error = test_err), aes(k, error)) +
  geom_line(colour="#e74c3c", linewidth=1.2) +
  geom_point(size=3, colour="#e74c3c") +
  geom_vline(xintercept=ks[which.min(test_err)], linetype="dashed", colour="#888") +
  scale_x_log10(breaks=ks) +
  labs(title="kNN test error vs k (SWP dataset)",
       subtitle=paste0("Optimal k = ", ks[which.min(test_err)]),
       x="k (log scale)", y="Test error rate") +
  theme_minimal()
```

```{r}
#| label: method-comparison
logit    <- glm(SWP ~ Age + CTW + JobDemands + JobControl + JobSupport + MentalWB,
                data=train, family=binomial)
k_best   <- ks[which.min(test_err)]
knn_pred <- knn(X_tr_s, X_te, cl=y_tr_s, k=k_best)

cat("Test accuracy:\n")
cat("  Null (always SWP=1):    ", round(mean(test$SWP==1)*100,1), "%\n")
cat("  Logistic regression:    ", round(mean(ifelse(predict(logit,test,type="response")>0.5,1,0)==test$SWP)*100,1), "%\n")
cat("  LDA:                    ", round(mean(lda_pred$class==test$SWP)*100,1), "%\n")
cat("  kNN (k=", k_best, "):          ", round(mean(knn_pred==test$SWP)*100,1), "%\n")
```

---

# Guided Practice

::: {.webr-loading-note}
**R is loading in the background** — exercises below will be interactive once ready.
:::

```{webr-r}
#| label: guided-lda
lda_fit <- lda(SWP ~ Age + CTW + JobDemands + JobControl + JobSupport + MentalWB,
               data = train)
cat("Class priors:\n"); print(round(lda_fit$prior, 4))
cat("\nClass means:\n");  print(round(lda_fit$means, 3))

# Which predictor differs most between SWP=0 and SWP=1?
diffs <- abs(lda_fit$means[2,] - lda_fit$means[1,])
cat("\nMean difference |SWP=1 - SWP=0|:\n")
print(round(sort(diffs, decreasing=TRUE), 3))
```

```{webr-r}
#| label: guided-knn
pred_vars <- c("Age","CTW","JobDemands","JobControl","JobSupport","MentalWB")
mu  <- colMeans(train[, pred_vars])
sds <- apply(train[, pred_vars], 2, sd)
X_tr <- scale(train[, pred_vars], center=mu, scale=sds)
X_te <- scale(test[, pred_vars],  center=mu, scale=sds)

for (k in c(5, 50, 200)) {
  pred <- knn(X_tr, X_te, cl=train$SWP, k=k)
  cm   <- table(Predicted=pred, Actual=test$SWP)
  sens <- if (nrow(cm)==2) cm[1,1]/sum(cm[,1]) else 0
  cat(sprintf("k=%3d  Acc=%.1f%%  Sens(SWP=0)=%.1f%%\n",
              k, mean(pred==test$SWP)*100, sens*100))
}
# TASK: Why does larger k hurt sensitivity for the minority class?
```

::: {.callout-note}
## Further reading
James et al. (2021). *An Introduction to Statistical Learning*. — Chapter 4.4 (LDA), Chapter 3.5 (kNN)
:::

