---
title: "R1 — OLS Review & Diagnostics"
engine: knitr
---

```{=html}
<script>document.body.classList.add("regression");</script>
```

```{r}
#| context: setup
#| include: false
library(ggplot2)
library(dplyr)
nhanes <- readRDS(here::here("data/nhanes.rds"))
```

```{webr-r}
#| context: setup
#| include: false
library(ggplot2)
library(dplyr)
nhanes <- readRDS("https://johlobs.github.io/STA230-App/data/nhanes.rds")
```

# Concept: OLS as the Baseline

## The OLS estimator

For $Y = \mathbf{X}\boldsymbol{\beta} + \boldsymbol{\varepsilon}$, OLS minimises the residual sum of squares:

$$\hat{\boldsymbol{\beta}} = \arg\min_{\boldsymbol{\beta}} \|\mathbf{y} - \mathbf{X}\boldsymbol{\beta}\|^2 = (\mathbf{X}^\top\mathbf{X})^{-1}\mathbf{X}^\top\mathbf{y}$$

**Properties (when assumptions hold):**

- **Unbiased**: $\mathbb{E}[\hat{\boldsymbol{\beta}}] = \boldsymbol{\beta}$
- **BLUE**: Best Linear Unbiased Estimator (Gauss-Markov)
- **Variance**: $\text{Var}(\hat{\boldsymbol{\beta}}) = \sigma^2 (\mathbf{X}^\top\mathbf{X})^{-1}$

## When OLS struggles

| Problem | Symptom | Consequence |
|---------|---------|-------------|
| $p \approx n$ | Many predictors, small sample | High variance, unstable $\hat{\beta}$ |
| $p > n$ | More predictors than observations | $(\mathbf{X}^\top\mathbf{X})$ singular — OLS undefined |
| Multicollinearity | Predictors correlated | Inflated SE, unreliable coefficients |
| Irrelevant predictors | Noise variables included | Higher variance, worse prediction |

## Assumptions and diagnostics

| Plot | What to look for |
|------|-----------------|
| Residuals vs fitted | Random scatter (linearity + homoskedasticity) |
| QQ-plot | Points on diagonal (normality) |
| Scale-location | Flat line (constant variance) |
| Residuals vs leverage | No high-leverage + high-residual outliers |

---

# Quiz

```{=html}
<div class="sta230-quiz" data-question="OLS is BLUE — Best Linear Unbiased Estimator. What does 'best' mean here?">
  <div class="q-opt" data-correct="false">Lowest test MSE among all estimators</div>
  <div class="q-opt" data-correct="true">Lowest variance among all linear unbiased estimators</div>
  <div class="q-opt" data-correct="false">Lowest bias among all estimators</div>
  <div class="q-opt" data-correct="false">Highest R² on training data</div>
  <div class="q-exp">Gauss-Markov: among estimators of the form β̂ = Cy (linear in Y) that are unbiased (E[β̂] = β), OLS has minimum variance. But 'best' is restricted to this class. A biased estimator (Ridge, LASSO) can have lower MSE if the variance reduction outweighs the bias introduced — exactly the bias-variance tradeoff.</div>
</div>

<div class="sta230-quiz" data-question="You add 10 irrelevant noise predictors to an OLS model. What happens to training R² and test MSE?">
  <div class="q-opt" data-correct="false">Both training R² and test MSE decrease</div>
  <div class="q-opt" data-correct="false">Training R² decreases, test MSE increases</div>
  <div class="q-opt" data-correct="true">Training R² stays the same or increases; test MSE increases</div>
  <div class="q-opt" data-correct="false">Both stay the same since the predictors are irrelevant</div>
  <div class="q-exp">Adding any predictor (even pure noise) can only increase or maintain training R² — OLS finds a non-zero coefficient that fits the training noise. On test data those noise coefficients hurt. Training fit improves, generalisation worsens — the classic overfitting pattern.</div>
</div>

<div class="sta230-quiz" data-question="A residual vs fitted plot shows a U-shaped pattern. What assumption is violated?">
  <div class="q-opt" data-correct="false">Normality of residuals</div>
  <div class="q-opt" data-correct="false">Independence of observations</div>
  <div class="q-opt" data-correct="true">Linearity — the true relationship is non-linear</div>
  <div class="q-opt" data-correct="false">Homoskedasticity — variance changes with fitted values</div>
  <div class="q-exp">A U-shape (systematic curve) in residuals vs fitted means the model consistently over- or underpredicts in certain ranges — a sign of non-linearity. A funnel shape (widening spread) would indicate heteroskedasticity instead. Both can appear together.</div>
</div>
```

---

# Worked Example: Predicting Systolic Blood Pressure (NHANES)

The dataset is NHANES 2015–2016 (n = 5 005 adults). Outcome: mean systolic blood pressure (SBP, mmHg). Predictors: age, sex, BMI, waist circumference, total cholesterol, diabetes, smoking status.

```{r}
#| label: ols-summary
fit <- lm(SBP ~ Age + Sex + BMI + Waist + Cholesterol + factor(Diabetes) + factor(Smoker),
          data = nhanes)
summary(fit)
```

```{r}
#| label: ols-diagnostics
#| fig-width: 10
#| fig-height: 5
par(mfrow = c(1, 2))
plot(fit, which = 1, main = "Residuals vs Fitted")
plot(fit, which = 2, main = "Normal Q-Q")
par(mfrow = c(1, 1))
```

```{r}
#| label: ols-train-test
set.seed(42)
train_idx <- sample(nrow(nhanes), round(0.7 * nrow(nhanes)))
train <- nhanes[train_idx, ]
test  <- nhanes[-train_idx, ]

fit_tr <- lm(SBP ~ Age + Sex + BMI + Waist + Cholesterol + factor(Diabetes) + factor(Smoker),
             data = train)

rmspe_train <- sqrt(mean(residuals(fit_tr)^2))
rmspe_test  <- sqrt(mean((test$SBP - predict(fit_tr, newdata = test))^2, na.rm = TRUE))

cat("Training RMSPE:", round(rmspe_train, 2), "mmHg\n")
cat("Test RMSPE:    ", round(rmspe_test,  2), "mmHg\n")
```

---

# Guided Practice

::: {.webr-loading-note}
**R is loading in the background** — the exercises below will be interactive once the status bar disappears.
:::

```{webr-r}
#| label: guided-explore
# TASK 1: Explore the dataset
cat("Dimensions:", nrow(nhanes), "x", ncol(nhanes), "\n")
summary(nhanes[, c("SBP", "Age", "BMI", "Waist", "Cholesterol")])
```

```{webr-r}
#| label: guided-ols
# TASK 2: Fit OLS of SBP on age, sex, and BMI
fit_small <- lm(SBP ~ Age + Sex + BMI, data = nhanes)
summary(fit_small)

# TASK 3: Add Waist, Cholesterol, Diabetes, Smoker — does R² improve much?
fit_full <- lm(SBP ~ Age + Sex + BMI + Waist + Cholesterol +
                 factor(Diabetes) + factor(Smoker), data = nhanes)

cat("\nSmall model R²:", round(summary(fit_small)$r.squared, 4), "\n")
cat("Full model R²: ", round(summary(fit_full)$r.squared, 4), "\n")
```

```{webr-r}
#| label: guided-split
# TASK 4: 70/30 train/test split — compare RMSPE for small vs full model
set.seed(42)
tr <- sample(nrow(nhanes), round(0.7 * nrow(nhanes)))
te <- setdiff(seq_len(nrow(nhanes)), tr)

fit_s <- lm(SBP ~ Age + Sex + BMI, data = nhanes[tr, ])
fit_f <- lm(SBP ~ Age + Sex + BMI + Waist + Cholesterol +
              factor(Diabetes) + factor(Smoker), data = nhanes[tr, ])

rmspe <- function(m, newdat)
  sqrt(mean((newdat$SBP - predict(m, newdata = newdat))^2, na.rm = TRUE))

cat("Small model — Train RMSPE:", round(sqrt(mean(residuals(fit_s)^2)), 2),
    " Test RMSPE:", round(rmspe(fit_s, nhanes[te, ]), 2), "\n")
cat("Full model  — Train RMSPE:", round(sqrt(mean(residuals(fit_f)^2)), 2),
    " Test RMSPE:", round(rmspe(fit_f, nhanes[te, ]), 2), "\n")
```

::: {.callout-tip collapse="true"}
## What to look for
- Does the full model have lower training RMSPE than the small model?
- Does it also have lower *test* RMSPE? How large is the optimism gap?
- Clinically: how precise are these predictions in mmHg? Is the model useful?
:::

---

# Retrieval Challenge

```{webr-r}
#| label: challenge
# Q1: Interpret the Age coefficient
fit <- lm(SBP ~ Age + Sex + BMI + Waist + Cholesterol +
            factor(Diabetes) + factor(Smoker), data = nhanes)
coef_age <- coef(fit)["Age"]
cat("Age coefficient:", round(coef_age, 3),
    "mmHg per year (holding other predictors constant)\n")

# Q2: Is BMI or Waist more strongly associated with SBP?
# (Compare t-statistics, not raw coefficients — different scales!)
s <- summary(fit)$coefficients
cat("\nBMI t-stat:   ", round(s["BMI",   "t value"], 2), "\n")
cat("Waist t-stat: ", round(s["Waist", "t value"], 2), "\n")
cat("Note: BMI and Waist are correlated — multicollinearity may affect both\n")

# Q3: Compute VIF manually for BMI
xj <- nhanes$BMI
others <- nhanes[, c("Age","Waist","Cholesterol")]
others$Sex <- as.numeric(nhanes$Sex == "Male")
r2_bmi <- summary(lm(xj ~ ., data = others))$r.squared
cat("\nVIF for BMI:", round(1 / (1 - r2_bmi), 2), "\n")
```

---

# Why-Questions

::: {.callout-caution collapse="true"}
## Q: "Both BMI and waist circumference are in the model. Is that OK?"

Both measure adiposity and are strongly correlated — this is multicollinearity. Including both inflates their standard errors, making individual coefficients less reliable. Options: (1) include only one (waist is often preferred as it captures central adiposity better), (2) use PCA to create a composite, (3) use Ridge/LASSO which handle multicollinearity automatically. For prediction, multicollinearity is less of a problem than for inference.
:::

::: {.callout-caution collapse="true"}
## Q: "My RMSPE is 17 mmHg. Is that a good model?"

It depends on context. A clinically meaningful difference in SBP is typically 5–10 mmHg. An RMSPE of 17 mmHg means individual predictions are off by ~17 mmHg on average — likely too imprecise for individual-level clinical decisions. However, the model may still be useful for population-level inference about which factors are associated with SBP. Precision and interpretability are separate goals.
:::

::: {.callout-note}
## Further reading
James et al. (2021). *An Introduction to Statistical Learning*. — Chapter 3
:::
