---
title: "R2 — Variable Selection"
engine: knitr
---

```{=html}
<script>document.body.classList.add("regression");</script>
```

```{r}
#| context: setup
#| include: false
library(ggplot2)
library(dplyr)
nhanes <- readRDS(here::here("data/nhanes.rds"))
nhanes_cc <- nhanes |> filter(!is.na(Diabetes), !is.na(Smoker))
```

```{webr-r}
#| context: setup
#| include: false
library(ggplot2)
library(dplyr)
nhanes <- readRDS("https://johlobs.github.io/STA230-App/data/nhanes.rds")
nhanes_cc <- nhanes |> filter(!is.na(Diabetes), !is.na(Smoker))
```

# Concept: Choosing Which Predictors to Include

## Why not include everything?

Adding predictors always reduces training error but increases variance. With $p$ predictors, OLS has $p+1$ parameters to estimate. When $p$ is large:

- Variance of $\hat{\beta}$ increases
- Test MSE may be higher than a model with fewer predictors
- Interpretation becomes harder

Variable selection finds the subset of predictors that minimises test error.

## Best subset selection

Fit all $2^p$ possible models, evaluate each by CV or an information criterion, pick the best. Exact but computationally infeasible for large $p$ — for $p=20$ there are over one million models.

## Forward stepwise selection

1. Start with the null model (intercept only)
2. At each step, add the predictor that most improves fit
3. Stop when adding more predictors no longer helps (by CV or AIC)

Much faster than best subset ($O(p^2)$ models), but not guaranteed to find the globally best subset.

## Backward stepwise selection

Start with the full model, remove the least useful predictor at each step. Requires $n > p$ to fit the full model.

## Criteria for stopping

| Criterion | Formula | Penalises |
|-----------|---------|-----------|
| AIC | $-2\ell + 2p$ | Predictive accuracy |
| BIC | $-2\ell + p\log(n)$ | Harder penalty, simpler models |
| Adjusted R² | $1 - \frac{\text{RSS}/(n-p-1)}{\text{TSS}/(n-1)}$ | Extra parameters |
| CV error | Estimated test MSE | Directly estimates generalisation |

---

# Quiz

```{=html}
<div class="sta230-quiz" data-question="Forward stepwise selection adds predictors one at a time. What is the main limitation compared to best subset selection?">
  <div class="q-opt" data-correct="false">It is slower than best subset selection</div>
  <div class="q-opt" data-correct="true">It is greedy — it cannot undo earlier choices, so it may miss the globally optimal subset</div>
  <div class="q-opt" data-correct="false">It requires n > p to work</div>
  <div class="q-opt" data-correct="false">It always selects more predictors than best subset</div>
  <div class="q-exp">Forward stepwise is greedy: each step picks the best predictor to add given what's already in the model. But a predictor that looks unhelpful early may become important once other predictors are added. Best subset avoids this by exhaustively trying all combinations, but is only feasible for small p (≤20 or so). In practice, LASSO is often preferred over both for large p.</div>
</div>

<div class="sta230-quiz" data-question="BIC penalises complexity more harshly than AIC for large n. Why?">
  <div class="q-opt" data-correct="false">BIC uses the test set; AIC uses the training set</div>
  <div class="q-opt" data-correct="true">BIC's penalty per parameter is log(n) vs 2 for AIC — for n > 8, log(n) > 2</div>
  <div class="q-opt" data-correct="false">BIC corrects for the optimism of training R²; AIC does not</div>
  <div class="q-opt" data-correct="false">AIC is only valid for linear models; BIC works for all models</div>
  <div class="q-exp">AIC penalty = 2p, BIC penalty = p·log(n). For n=5000 (like NHANES), log(n)≈8.5 — BIC penalises each extra predictor more than 4 times as hard as AIC. This makes BIC select much sparser models. BIC is consistent (selects true model as n→∞ under certain conditions); AIC aims for predictive accuracy and may overfit slightly.</div>
</div>

<div class="sta230-quiz" data-question="You compare models using adjusted R². A model with p=6 has adj-R²=0.42 and a model with p=3 has adj-R²=0.41. Which do you prefer?">
  <div class="q-opt" data-correct="false">The p=6 model — higher adjusted R² is always better</div>
  <div class="q-opt" data-correct="false">The p=3 model — fewer predictors always wins</div>
  <div class="q-opt" data-correct="true">The p=6 model is slightly better by this criterion, but the difference is small — verify with CV</div>
  <div class="q-opt" data-correct="false">Cannot compare — adjusted R² requires the same number of predictors</div>
  <div class="q-exp">Adjusted R² accounts for the number of predictors, so comparing across model sizes is valid. A difference of 0.01 is small and may not reflect a meaningful improvement in test performance. Cross-validation would give a more direct estimate of generalisation. In practice, use adjusted R² as a quick screening tool, CV for final model selection.</div>
</div>
```

---

# Worked Example

```{r}
#| label: forward-stepwise
# Forward stepwise selection with AIC using step()
fit_null <- lm(SBP ~ 1, data = nhanes_cc)
fit_full <- lm(SBP ~ Age + Sex + BMI + Waist + Cholesterol +
                 factor(Diabetes) + factor(Smoker), data = nhanes_cc)

fit_forward <- step(fit_null,
                    scope = list(lower = fit_null, upper = fit_full),
                    direction = "forward", trace = FALSE)

cat("Selected model formula:\n")
print(formula(fit_forward))
cat("\nAIC:", round(AIC(fit_forward), 2), "\n")
cat("BIC:", round(BIC(fit_forward), 2), "\n")
```

```{r}
#| label: backward-stepwise
fit_backward <- step(fit_full, direction = "backward", trace = FALSE)
cat("Backward stepwise selected formula:\n")
print(formula(fit_backward))

# Do forward and backward agree?
cat("\nForward and backward select the same model:",
    identical(sort(names(coef(fit_forward))),
              sort(names(coef(fit_backward)))), "\n")
```

```{r}
#| label: cv-model-comparison
#| fig-width: 8
#| fig-height: 4
# Compare test RMSPE for models with 1–6 predictors (forward order)
set.seed(42)
k <- 10
folds <- sample(rep(1:k, length.out = nrow(nhanes_cc)))

predictor_sets <- list(
  "Age only"              = "SBP ~ Age",
  "Age + Sex"             = "SBP ~ Age + Sex",
  "Age + Sex + BMI"       = "SBP ~ Age + Sex + BMI",
  "Age + Sex + BMI + Waist" = "SBP ~ Age + Sex + BMI + Waist",
  "Age + Sex + BMI + Waist + Chol" = "SBP ~ Age + Sex + BMI + Waist + Cholesterol",
  "Full model"            = "SBP ~ Age + Sex + BMI + Waist + Cholesterol + factor(Diabetes) + factor(Smoker)"
)

cv_rmspe <- sapply(predictor_sets, function(fmla) {
  sqrt(mean(sapply(1:k, function(j) {
    tr <- nhanes_cc[folds != j, ]
    te <- nhanes_cc[folds == j, ]
    fit <- lm(as.formula(fmla), data = tr)
    mean((te$SBP - predict(fit, newdata = te))^2, na.rm = TRUE)
  })))
})

res <- data.frame(Model = factor(names(cv_rmspe), levels = names(cv_rmspe)),
                  RMSPE = cv_rmspe)

ggplot(res, aes(x = Model, y = RMSPE)) +
  geom_col(fill = "#2980b9", alpha = 0.8) +
  geom_text(aes(label = round(RMSPE, 2)), vjust = -0.4, size = 3.5) +
  labs(title = "10-fold CV RMSPE by model size (NHANES, outcome = SBP)",
       x = NULL, y = "RMSPE (mmHg)") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 25, hjust = 1))
```

---

# Guided Practice

::: {.webr-loading-note}
**R is loading in the background** — exercises below will be interactive once ready.
:::

```{webr-r}
#| label: guided-forward
# TASK 1: Run forward stepwise selection using step()
nhanes_cc <- nhanes |> filter(!is.na(Diabetes), !is.na(Smoker))

fit_null <- lm(SBP ~ 1, data = nhanes_cc)
fit_full <- lm(SBP ~ Age + Sex + BMI + Waist + Cholesterol +
                 factor(Diabetes) + factor(Smoker), data = nhanes_cc)

fit_fwd <- step(fit_null,
                scope = list(lower = fit_null, upper = fit_full),
                direction = "forward", trace = TRUE)  # trace=TRUE shows steps

cat("\nFinal model:\n")
print(formula(fit_fwd))
```

```{webr-r}
#| label: guided-cv
# TASK 2: Manually compare 3-predictor vs full model via 5-fold CV
set.seed(42)
nhanes_cc <- nhanes |> filter(!is.na(Diabetes), !is.na(Smoker))
k <- 5
folds <- sample(rep(1:k, length.out = nrow(nhanes_cc)))

cv_rmspe <- function(fmla) {
  sqrt(mean(sapply(1:k, function(j) {
    tr <- nhanes_cc[folds != j, ]
    te <- nhanes_cc[folds == j, ]
    fit <- lm(as.formula(fmla), data = tr)
    mean((te$SBP - predict(fit, newdata = te))^2, na.rm = TRUE)
  })))
}

cat("3-predictor CV RMSPE:", round(cv_rmspe("SBP ~ Age + Sex + BMI"), 3), "mmHg\n")
cat("Full model CV RMSPE: ", round(cv_rmspe("SBP ~ Age + Sex + BMI + Waist + Cholesterol + factor(Diabetes) + factor(Smoker)"), 3), "mmHg\n")

# TASK 3: Which model would you choose? Does adding Waist and Cholesterol meaningfully help?
```

::: {.callout-tip collapse="true"}
## Hints
- Look at the AIC trace from `step()` — it shows which predictor was added at each step and the AIC change
- A RMSPE difference of <0.5 mmHg between models is probably not clinically meaningful
- The 1-SE rule: if the simpler model is within 1 SE of the best, prefer it
:::

---

# Retrieval Challenge

```{webr-r}
#| label: challenge
# Q1: How many models does best subset selection fit for p=7 predictors?
cat("Best subset models for p=7:", 2^7, "\n")

# Q2: Compute AIC and BIC manually for a fitted model
nhanes_cc <- nhanes |> filter(!is.na(Diabetes), !is.na(Smoker))
fit <- lm(SBP ~ Age + Sex + BMI + Waist, data = nhanes_cc)
n <- nrow(nhanes_cc)
p <- length(coef(fit))
ll <- logLik(fit)

cat("\nLog-likelihood:", round(as.numeric(ll), 2), "\n")
cat("AIC (manual):", round(-2 * as.numeric(ll) + 2 * p, 2),
    "| AIC():", round(AIC(fit), 2), "\n")
cat("BIC (manual):", round(-2 * as.numeric(ll) + log(n) * p, 2),
    "| BIC():", round(BIC(fit), 2), "\n")
cat("BIC penalty per param:", round(log(n), 2), "(vs AIC: 2)\n")
```

::: {.callout-note}
## Further reading
James et al. (2021). *An Introduction to Statistical Learning*. — Chapter 6.1 (Subset selection)
:::
