---
title: "R3 — Ridge Regression"
engine: knitr
---

```{=html}
<script>document.body.classList.add("regression");</script>
```

```{r}
#| context: setup
#| include: false
library(ggplot2)
library(dplyr)
library(glmnet)
nhanes <- readRDS(here::here("data/nhanes.rds"))
nhanes_cc <- nhanes |> filter(!is.na(Diabetes), !is.na(Smoker))
X <- model.matrix(SBP ~ Age + Sex + BMI + Waist + Cholesterol +
                    factor(Diabetes) + factor(Smoker), data = nhanes_cc)[, -1]
y <- nhanes_cc$SBP
```

```{webr-r}
#| context: setup
#| include: false
library(ggplot2)
library(dplyr)
library(glmnet)
nhanes <- readRDS("https://johlobs.github.io/STA230-App/data/nhanes.rds")
nhanes_cc <- nhanes |> filter(!is.na(Diabetes), !is.na(Smoker))
X <- model.matrix(SBP ~ Age + Sex + BMI + Waist + Cholesterol +
                    factor(Diabetes) + factor(Smoker), data = nhanes_cc)[, -1]
y <- nhanes_cc$SBP
```

# Concept: Regularisation via L2 Penalty

## The ridge objective

Ridge regression adds an $L_2$ penalty to the OLS loss:

$$\hat{\boldsymbol{\beta}}^{\text{ridge}} = \arg\min_{\boldsymbol{\beta}} \left\{ \sum_{i=1}^n (y_i - \mathbf{x}_i^\top\boldsymbol{\beta})^2 + \lambda \sum_{j=1}^p \beta_j^2 \right\}$$

The penalty $\lambda \sum \beta_j^2$ shrinks all coefficients toward zero. The **tuning parameter** $\lambda \geq 0$ controls the amount of shrinkage:

- $\lambda = 0$: OLS (no shrinkage)
- $\lambda \to \infty$: all $\hat{\beta}_j \to 0$

## Closed-form solution

$$\hat{\boldsymbol{\beta}}^{\text{ridge}} = (\mathbf{X}^\top\mathbf{X} + \lambda\mathbf{I})^{-1}\mathbf{X}^\top\mathbf{y}$$

Adding $\lambda\mathbf{I}$ makes the matrix invertible even when $\mathbf{X}^\top\mathbf{X}$ is singular or near-singular — this is why ridge works when $p \geq n$.

## Bias-variance effect

| $\lambda$ | Bias | Variance | Test MSE |
|-----------|------|----------|---------|
| 0 (OLS)   | 0    | High     | May be high |
| Small     | Low  | Lower    | Often lower |
| Large     | High | Very low | May be high |

**Key insight:** Ridge introduces bias but reduces variance. The optimal $\lambda$ minimises test MSE — found via cross-validation.

## Important: standardise predictors

Ridge shrinks coefficients, but the amount of shrinkage depends on predictor scale. **Always standardise** predictors before applying ridge (and LASSO). `glmnet()` does this automatically by default.

---

# Quiz

```{=html}
<div class="sta230-quiz" data-question="Why does ridge regression shrink coefficients but never set them exactly to zero?">
  <div class="q-opt" data-correct="false">The L2 penalty is simply not large enough mathematically to force any coefficient to exactly zero</div>
  <div class="q-opt" data-correct="true">The L2 gradient 2λβ → 0 as β → 0, so the optimal β is always small but never exactly zero</div>
  <div class="q-opt" data-correct="false">A smooth L2 penalty has no kink at zero, unlike L1, so it cannot produce exact sparse solutions</div>
  <div class="q-opt" data-correct="false">Ridge zeros out large coefficients while leaving small coefficients unchanged by the L2 penalty</div>
  <div class="q-exp">The L2 penalty contributes a gradient of 2λβ_j to each coefficient. As β_j gets small, the gradient also gets small — so the optimum is at some small but non-zero value. In contrast, the L1 penalty (LASSO) has a constant subgradient of λ near zero, which forces the optimum to exactly zero for small enough coefficients. This is why LASSO performs variable selection and ridge does not.</div>
</div>

<div class="sta230-quiz" data-question="Ridge is preferred over LASSO when you believe...">
  <div class="q-opt" data-correct="false">Only a few predictors truly affect the outcome</div>
  <div class="q-opt" data-correct="true">Many predictors each have a small true effect</div>
  <div class="q-opt" data-correct="false">Predictors are independent of each other</div>
  <div class="q-opt" data-correct="false">You need an interpretable sparse solution</div>
  <div class="q-exp">Ridge is best when the true signal is spread across many predictors — it shrinks all coefficients moderately. LASSO is best when only a few predictors truly matter — it zeros out the rest. In GWAS (genetic association studies), for example, many thousands of SNPs each have tiny effects — ridge is often preferred. For clinical risk scores, LASSO is preferred for sparsity.</div>
</div>

<div class="sta230-quiz" data-question="You fit ridge regression on unstandardised predictors where Age is in years (0–80) and Cholesterol is in mg/dL (100–300). What problem arises?">
  <div class="q-opt" data-correct="false">Ridge breaks down entirely when predictors have different scales — the algorithm cannot produce a solution</div>
  <div class="q-opt" data-correct="true">Large-scale predictors have bigger β² and face more penalty shrinkage than small-scale ones</div>
  <div class="q-opt" data-correct="false">The algorithm fails to converge numerically when predictors span very different measurement scales</div>
  <div class="q-opt" data-correct="false">No problem — the ridge penalty automatically adjusts for predictor scale before applying shrinkage</div>
  <div class="q-exp">The L2 penalty is λΣβ_j². A predictor measured in large units (like Cholesterol in mg/dL) will have a small coefficient even for a strong association — so its coefficient gets penalised less. A predictor in small units has a large coefficient even for a weak effect — and gets penalised more. Standardising removes this scale dependence. glmnet() standardises by default.</div>
</div>
```

---

# Worked Example

```{r}
#| label: ridge-fit
set.seed(42)
# glmnet with alpha=0 is ridge
# By default standardises predictors and fits over a lambda grid
ridge_fit <- glmnet(X, y, alpha = 0)

# Coefficient paths
plot(ridge_fit, xvar = "lambda", label = TRUE,
     main = "Ridge coefficient paths (NHANES → SBP)")
```

```{r}
#| label: ridge-cv
# 10-fold CV to find optimal lambda
set.seed(42)
cv_ridge <- cv.glmnet(X, y, alpha = 0, nfolds = 10)
plot(cv_ridge)

cat("lambda.min:", round(cv_ridge$lambda.min, 4), "\n")
cat("lambda.1se:", round(cv_ridge$lambda.1se, 4), "\n")
cat("\nCV RMSPE at lambda.min:", round(sqrt(min(cv_ridge$cvm)), 3), "mmHg\n")
```

```{r}
#| label: ridge-vs-ols
# Compare ridge vs OLS on a train/test split
set.seed(42)
tr_idx <- sample(nrow(nhanes_cc), round(0.7 * nrow(nhanes_cc)))
X_tr <- X[tr_idx, ];   y_tr <- y[tr_idx]
X_te <- X[-tr_idx, ];  y_te <- y[-tr_idx]

ols_fit   <- lm(y_tr ~ X_tr)
ridge_cv  <- cv.glmnet(X_tr, y_tr, alpha = 0, nfolds = 10)

ols_rmspe   <- sqrt(mean((y_te - cbind(1, X_te) %*% coef(ols_fit))^2))
ridge_rmspe <- sqrt(mean((y_te - predict(ridge_cv, newx = X_te, s = "lambda.min"))^2))

cat("OLS test RMSPE:   ", round(ols_rmspe,   3), "mmHg\n")
cat("Ridge test RMSPE: ", round(ridge_rmspe, 3), "mmHg\n")
```

---

# Guided Practice

::: {.webr-loading-note}
**R is loading in the background** — exercises below will be interactive once ready.
:::

```{webr-r}
#| label: guided-ridge-path
# TASK 1: Fit ridge and inspect coefficient paths
ridge_fit <- glmnet(X, y, alpha = 0)

# How do coefficients change with lambda?
# Pick a few lambda values and inspect coefficients
lambdas <- c(0.1, 1, 10, 100, 1000)
coef_mat <- sapply(lambdas, function(l)
  as.vector(coef(ridge_fit, s = l)))
rownames(coef_mat) <- rownames(coef(ridge_fit, s = 1))
colnames(coef_mat) <- paste0("λ=", lambdas)
round(coef_mat, 4)
```

```{webr-r}
#| label: guided-ridge-cv
# TASK 2: Run 10-fold CV and find optimal lambda
set.seed(42)
cv_ridge <- cv.glmnet(X, y, alpha = 0, nfolds = 10)

cat("lambda.min:", round(cv_ridge$lambda.min, 4), "\n")
cat("lambda.1se:", round(cv_ridge$lambda.1se, 4), "\n")
cat("CV RMSPE at lambda.min:", round(sqrt(min(cv_ridge$cvm)), 3), "mmHg\n")

# TASK 3: Extract and interpret the ridge coefficients at lambda.min
coef_ridge <- coef(cv_ridge, s = "lambda.min")
print(round(coef_ridge, 4))

# Compare to OLS coefficients
ols <- lm(y ~ X)
cat("\nOLS coefficients:\n")
print(round(coef(ols), 4))
```

::: {.callout-tip collapse="true"}
## What to compare
- Are the ridge coefficients larger or smaller than OLS?
- Which predictor is shrunk the most? Does that make sense?
- Does the ridge RMSPE meaningfully beat OLS here? (With n=4000+ and p=7, shrinkage may help little)
:::

---

# Why-Questions

::: {.callout-caution collapse="true"}
## Q: "Ridge keeps all predictors. How is that useful for interpretation?"

Ridge is primarily a **prediction** method, not an inference method. If interpretation matters, report OLS coefficients (with SE and p-values) from a carefully selected model. Use ridge when your goal is accurate prediction with many correlated predictors — for example, predicting future SBP from a panel of biomarkers. The stability of ridge coefficients under collinearity is a feature, not a bug, for prediction purposes.
:::

::: {.callout-caution collapse="true"}
## Q: "Does it matter if I standardise before or inside glmnet?"

`glmnet()` standardises internally by default (standardize=TRUE), fits on the standardised scale, then returns coefficients on the *original* scale. This means you can pass raw data and interpret coefficients directly. If you standardise manually first, set standardize=FALSE. Never standardise the outcome y for ridge/LASSO regression — only the predictors X.
:::

::: {.callout-note}
## Further reading
James et al. (2021). *An Introduction to Statistical Learning*. — Chapter 6.2
:::
