---
title: "R4 — LASSO"
engine: knitr
---

```{=html}
<script>document.body.classList.add("regression");</script>
```

```{r}
#| context: setup
#| include: false
library(ggplot2)
library(dplyr)
library(glmnet)
nhanes <- readRDS(here::here("data/nhanes.rds"))
nhanes_cc <- nhanes |> filter(!is.na(Diabetes), !is.na(Smoker))
X <- model.matrix(SBP ~ Age + Sex + BMI + Waist + Cholesterol +
                    factor(Diabetes) + factor(Smoker), data = nhanes_cc)[, -1]
y <- nhanes_cc$SBP
```

```{webr-r}
#| context: setup
#| include: false
library(ggplot2)
library(dplyr)
library(glmnet)
nhanes <- readRDS("https://johlobs.github.io/STA230-App/data/nhanes.rds")
nhanes_cc <- nhanes |> filter(!is.na(Diabetes), !is.na(Smoker))
X <- model.matrix(SBP ~ Age + Sex + BMI + Waist + Cholesterol +
                    factor(Diabetes) + factor(Smoker), data = nhanes_cc)[, -1]
y <- nhanes_cc$SBP
```

# Concept: Regularisation via L1 Penalty

## The LASSO objective

LASSO (Least Absolute Shrinkage and Selection Operator) uses an $L_1$ penalty:

$$\hat{\boldsymbol{\beta}}^{\text{lasso}} = \arg\min_{\boldsymbol{\beta}} \left\{ \sum_{i=1}^n (y_i - \mathbf{x}_i^\top\boldsymbol{\beta})^2 + \lambda \sum_{j=1}^p |\beta_j| \right\}$$

The key difference from ridge: the $L_1$ penalty produces **sparse solutions** — many $\hat{\beta}_j = 0$ exactly. LASSO does **variable selection** automatically.

## Why does L1 produce zeros?

Geometrically, the $L_1$ constraint region (a diamond in 2D) has corners on the axes. The OLS ellipsoidal contours are likely to first touch one of these corners, where one or more $\beta_j = 0$. The $L_2$ constraint (a circle) has no corners, so the solution is rarely exactly zero.

## LASSO vs Ridge

| | LASSO | Ridge |
|---|-------|-------|
| Penalty | $\lambda \sum |\beta_j|$ | $\lambda \sum \beta_j^2$ |
| Sparsity | Yes — sets coefficients to exactly 0 | No — shrinks toward 0 but not to 0 |
| Variable selection | Yes | No |
| Best when | Few predictors truly matter | Many predictors with small effects |
| Correlated predictors | Selects one arbitrarily | Shrinks all equally |

## Elastic Net

The **elastic net** combines both penalties:

$$\lambda \left[\alpha \sum |\beta_j| + (1-\alpha) \sum \beta_j^2 \right]$$

With $\alpha=1$ this is LASSO; $\alpha=0$ is ridge. Elastic net handles correlated predictors better than pure LASSO. In `glmnet()`, `alpha` controls the mix.

---

# Quiz

```{=html}
<div class="sta230-quiz" data-question="LASSO sets some coefficients exactly to zero. What practical advantage does this give?">
  <div class="q-opt" data-correct="false">It always produces a lower test MSE than ridge</div>
  <div class="q-opt" data-correct="true">It performs automatic variable selection — the non-zero predictors form an interpretable sparse model</div>
  <div class="q-opt" data-correct="false">It avoids the need for cross-validation to choose λ</div>
  <div class="q-opt" data-correct="false">It guarantees unbiased coefficient estimates for non-zero predictors</div>
  <div class="q-exp">LASSO's sparsity is its key practical advantage. When p is large (say 100 predictors), LASSO might select 8 non-zero coefficients — these form a concise, interpretable model. In biostatistics, this is often exactly what's needed: a clinical risk score with a small number of measurable variables. Note: LASSO coefficients are biased (shrunk toward zero), so they should not be used directly for inference without correction.</div>
</div>

<div class="sta230-quiz" data-question="cv.glmnet() returns lambda.min and lambda.1se. In practice, which is often preferred and why?">
  <div class="q-opt" data-correct="false">lambda.min — it minimises CV error so it is always the best choice</div>
  <div class="q-opt" data-correct="true">lambda.1se — it selects the simplest model whose CV error is within 1 SE of the minimum, giving a sparser and more robust model</div>
  <div class="q-opt" data-correct="false">lambda.min for prediction, lambda.1se for hypothesis testing</div>
  <div class="q-opt" data-correct="false">lambda.1se — it always has lower test MSE</div>
  <div class="q-exp">lambda.min gives the best average CV performance. lambda.1se uses the 1-SE rule: pick the largest λ (most regularisation, fewest non-zero predictors) whose CV error is within 1 standard error of the minimum. Since the CV errors are often not statistically different across a range of λ, the 1-SE model is sparser but statistically comparable — a good default for building parsimonious models.</div>
</div>

<div class="sta230-quiz" data-question="Two predictors X1 and X2 are perfectly correlated (r=1). What does LASSO do?">
  <div class="q-opt" data-correct="false">Includes both with equal coefficients</div>
  <div class="q-opt" data-correct="false">Excludes both since they are redundant</div>
  <div class="q-opt" data-correct="true">Arbitrarily selects one and sets the other to zero — the solution is not unique</div>
  <div class="q-opt" data-correct="false">Crashes — LASSO cannot handle perfectly correlated predictors</div>
  <div class="q-exp">With perfect correlation, the LASSO problem has many equivalent solutions (any split of the effect between X1 and X2 gives the same fit). LASSO arbitrarily picks one. Elastic net handles this better: it groups correlated predictors and shrinks them together, distributing the effect across the group. For highly correlated predictors (like BMI and waist circumference), elastic net is more stable than LASSO.</div>
</div>
```

---

# Worked Example

```{r}
#| label: lasso-path
lasso_fit <- glmnet(X, y, alpha = 1)
plot(lasso_fit, xvar = "lambda", label = TRUE,
     main = "LASSO coefficient paths (NHANES → SBP)")
```

```{r}
#| label: lasso-cv
set.seed(42)
cv_lasso <- cv.glmnet(X, y, alpha = 1, nfolds = 10)
plot(cv_lasso)

cat("lambda.min:", round(cv_lasso$lambda.min, 4), "\n")
cat("lambda.1se:", round(cv_lasso$lambda.1se, 4), "\n")

# Coefficients at lambda.1se
coef_1se <- coef(cv_lasso, s = "lambda.1se")
cat("\nNon-zero coefficients at lambda.1se:\n")
nonzero <- coef_1se[coef_1se[, 1] != 0, , drop = FALSE]
print(round(nonzero, 4))
cat("\nNumber of non-zero predictors:", nrow(nonzero) - 1, "\n")
```

```{r}
#| label: lasso-vs-ridge
# Compare LASSO, Ridge, and OLS on train/test
set.seed(42)
tr_idx <- sample(nrow(nhanes_cc), round(0.7 * nrow(nhanes_cc)))
X_tr <- X[tr_idx, ];  y_tr <- y[tr_idx]
X_te <- X[-tr_idx, ]; y_te <- y[-tr_idx]

cv_l <- cv.glmnet(X_tr, y_tr, alpha = 1, nfolds = 10)
cv_r <- cv.glmnet(X_tr, y_tr, alpha = 0, nfolds = 10)
ols  <- lm(y_tr ~ X_tr)

rmspe <- function(pred, ytrue) sqrt(mean((ytrue - pred)^2))

cat("OLS test RMSPE:   ", round(rmspe(cbind(1, X_te) %*% coef(ols), y_te), 3), "\n")
cat("Ridge test RMSPE: ", round(rmspe(predict(cv_r, X_te, s="lambda.min"), y_te), 3), "\n")
cat("LASSO test RMSPE: ", round(rmspe(predict(cv_l, X_te, s="lambda.min"), y_te), 3), "\n")
```

---

# Guided Practice

::: {.webr-loading-note}
**R is loading in the background** — exercises below will be interactive once ready.
:::

```{webr-r}
#| label: guided-lasso
# TASK 1: Fit LASSO with cv.glmnet and inspect results
set.seed(42)
cv_lasso <- cv.glmnet(X, y, alpha = 1, nfolds = 10)
cat("lambda.min:", round(cv_lasso$lambda.min, 4), "\n")
cat("lambda.1se:", round(cv_lasso$lambda.1se, 4), "\n")

# TASK 2: Which predictors are retained at lambda.1se?
coef_1se <- coef(cv_lasso, s = "lambda.1se")
cat("\nCoefficients at lambda.1se:\n")
print(round(coef_1se, 4))
```

```{webr-r}
#| label: guided-elastic
# TASK 3: Try elastic net (alpha = 0.5) — does it select different predictors?
set.seed(42)
cv_enet <- cv.glmnet(X, y, alpha = 0.5, nfolds = 10)
coef_enet <- coef(cv_enet, s = "lambda.1se")

cat("Elastic net non-zero predictors:\n")
print(round(coef_enet[coef_enet[,1] != 0, , drop=FALSE], 4))

# TASK 4: Compare test RMSPE for LASSO vs elastic net (70/30 split)
set.seed(42)
tr <- sample(nrow(nhanes_cc), round(0.7 * nrow(nhanes_cc)))
te <- setdiff(seq_len(nrow(nhanes_cc)), tr)
X_tr <- X[tr,]; y_tr <- y[tr]
X_te <- X[te,]; y_te <- y[te]

cv_l2 <- cv.glmnet(X_tr, y_tr, alpha = 1)
cv_e2 <- cv.glmnet(X_tr, y_tr, alpha = 0.5)
rmspe <- function(p, yt) sqrt(mean((yt - p)^2))

cat("\nLASSO test RMSPE:       ", round(rmspe(predict(cv_l2, X_te, s="lambda.min"), y_te), 3), "\n")
cat("Elastic net test RMSPE: ", round(rmspe(predict(cv_e2, X_te, s="lambda.min"), y_te), 3), "\n")
```

::: {.callout-tip collapse="true"}
## What to compare
- Does LASSO zero out BMI or Waist? They are correlated — LASSO may arbitrarily pick one
- Does elastic net handle this differently?
- Is the test RMSPE meaningfully different between methods?
:::

---

# Retrieval Challenge

```{webr-r}
#| label: challenge
# Simulation: LASSO for variable selection when only a few predictors matter
set.seed(99)
n <- 200; p <- 20
X_sim <- matrix(rnorm(n * p), n, p)
colnames(X_sim) <- paste0("X", 1:p)
# True model: only X2, X5, X11 matter
beta_true <- rep(0, p)
beta_true[c(2, 5, 11)] <- c(3, -2, 1.5)
y_sim <- X_sim %*% beta_true + rnorm(n, 0, 2)

# Q1: Fit LASSO — does it recover X2, X5, X11?
cv_sim <- cv.glmnet(X_sim, y_sim, alpha = 1)
coef_sim <- coef(cv_sim, s = "lambda.1se")
selected <- rownames(coef_sim)[coef_sim[,1] != 0]
cat("Selected predictors at lambda.1se:", paste(selected[-1], collapse=", "), "\n")
cat("True predictors: X2, X5, X11\n")

# Q2: Does lambda.min or lambda.1se recover the true model better?
coef_min <- coef(cv_sim, s = "lambda.min")
sel_min  <- rownames(coef_min)[coef_min[,1] != 0]
cat("\nSelected at lambda.min:", paste(sel_min[-1], collapse=", "), "\n")
```

::: {.callout-note collapse="true"}
## Model answer
With n=200 and p=20, LASSO should recover X2, X5, X11 reasonably well. lambda.1se tends to be more conservative (fewer predictors) and may miss weak signals (X11 with β=1.5). lambda.min includes more predictors and is more likely to recover all three but may also include false positives.
:::

::: {.callout-note}
## Further reading
James et al. (2021). *An Introduction to Statistical Learning*. — Chapter 6.2
:::
