---
title: "U1 — Principal Component Analysis"
engine: knitr
---

```{=html}
<script>document.body.classList.add("unsupervised");</script>
```

```{r}
#| context: setup
#| include: false
library(ggplot2)
library(dplyr)
nhanes <- readRDS(here::here("data/nhanes.rds"))
```

```{webr-r}
#| context: setup
#| include: false
library(ggplot2)
library(dplyr)
nhanes <- readRDS("https://johlobs.github.io/STA230-App/data/nhanes.rds")
```

# Concept: From STA220 to PCA

## You have seen this before

In STA220 you used **factor analysis** (FA) on the HAD scale — 14 items measuring depression and anxiety. FA found two latent factors: one loading on depression items, one on anxiety. Both factors were hidden dimensions that explained the correlations among items.

**PCA asks a related but different question:**

| | Factor Analysis (STA220) | PCA (STA230) |
|---|---|---|
| **Goal** | Find latent factors that *cause* correlations | Find directions that capture *maximum variance* |
| **Model** | $\mathbf{X} = \boldsymbol{\Lambda}\mathbf{f} + \boldsymbol{\varepsilon}$ (generative) | Geometric rotation of the data cloud |
| **Output** | Interpretable factors (depression, anxiety) | Principal components ordered by variance |
| **Use case** | Psychological measurement, construct validity | Compression, visualisation, multicollinearity |

Both methods reduce many correlated variables to a smaller set of dimensions. PCA is simpler and makes no distributional assumptions — it is purely about variance.

## The geometry of PCA

Given $n$ observations on $p$ variables, PCA finds a sequence of orthogonal directions (principal components) $\mathbf{z}_1, \mathbf{z}_2, \ldots$ such that:

$$\mathbf{z}_1 = \mathbf{X}\mathbf{v}_1, \quad \text{where } \mathbf{v}_1 = \arg\max_{\|\mathbf{v}\|=1} \text{Var}(\mathbf{X}\mathbf{v})$$

Each subsequent component captures the maximum remaining variance, orthogonal to all previous ones.

In practice: **standardise the variables** (subtract mean, divide by SD), then compute the eigendecomposition of the covariance (or correlation) matrix:

$$\mathbf{\Sigma} = \mathbf{V} \mathbf{D} \mathbf{V}^\top$$

- $\mathbf{V}$: eigenvectors = directions of principal components (loadings)
- $\mathbf{D}$: diagonal matrix of eigenvalues = variance explained by each component

**Proportion of variance explained** by component $k$: $\lambda_k / \sum_j \lambda_j$

## Reading PCA output

| Output | What it tells you |
|--------|-------------------|
| **Scree plot** | Eigenvalues in decreasing order — look for an "elbow" |
| **Loadings** | How much each original variable contributes to each PC |
| **Biplot** | Observations (points) and variables (arrows) in PC space |
| **Cumulative variance** | How many PCs needed to explain e.g. 80% of variance |

## Scaling matters

Always use `scale = TRUE` in `prcomp()`. Variables on different scales (age in years, cholesterol in mmol/L) would otherwise be dominated by whichever has the largest variance — purely an artefact of measurement units, not biology.

---

# Quiz

```{=html}
<div class="sta230-quiz" data-question="In STA220, factor analysis on the HAD scale found two factors (depression, anxiety). What is the main difference if you run PCA on the same items instead?">
  <div class="q-opt" data-correct="false">PCA would find more factors than FA because it uses all variance</div>
  <div class="q-opt" data-correct="true">PCA finds directions of maximum variance without assuming a generative model — the components may be similar but are interpreted geometrically, not as latent constructs</div>
  <div class="q-opt" data-correct="false">PCA and FA always give identical results on the same data</div>
  <div class="q-opt" data-correct="false">PCA requires a response variable Y; FA does not</div>
  <div class="q-exp">FA assumes X = Λf + ε — latent factors causally produce the observed items. PCA makes no such assumption: it finds orthogonal directions that explain the most variance. On HAD data the first two PCs would likely align with the depression/anxiety split, but a PC is not a "latent factor" — it is a linear combination that maximises explained variance.</div>
</div>

<div class="sta230-quiz" data-question="You run prcomp(X, scale = TRUE) on five health variables. The eigenvalues are: 2.4, 1.1, 0.7, 0.5, 0.3. How much variance does PC1 explain?">
  <div class="q-opt" data-correct="false">2.4%</div>
  <div class="q-opt" data-correct="false">24%</div>
  <div class="q-opt" data-correct="true">48% — eigenvalue 2.4 out of a total of 5.0 (= number of variables when scaled)</div>
  <div class="q-opt" data-correct="false">It cannot be determined without the raw data</div>
  <div class="q-exp">With scale = TRUE, the total variance equals the number of variables (each standardised variable has variance 1). Here total = 2.4 + 1.1 + 0.7 + 0.5 + 0.3 = 5.0. PC1 explains 2.4/5.0 = 48%. The first two PCs together explain (2.4 + 1.1)/5.0 = 70%.</div>
</div>

<div class="sta230-quiz" data-question="A biplot arrow for 'Waist' and an arrow for 'BMI' point in nearly the same direction. What does this indicate?">
  <div class="q-opt" data-correct="false">Waist and BMI are uncorrelated in this dataset</div>
  <div class="q-opt" data-correct="true">Waist and BMI are strongly positively correlated — they load similarly on the principal components</div>
  <div class="q-opt" data-correct="false">Waist causes BMI or vice versa</div>
  <div class="q-opt" data-correct="false">Both variables should be removed from the analysis</div>
  <div class="q-exp">In a biplot, arrows pointing in the same direction indicate variables that move together (are positively correlated). This is exactly what the OLS section identified as multicollinearity between Waist and BMI — PCA reveals this structure visually. One practical use of PCA is to replace correlated predictors with uncorrelated PCs before regression.</div>
</div>
```

---

# Worked Example: Health Variables in NHANES

We run PCA on five correlated health variables: Age, BMI, Waist circumference, total Cholesterol, and systolic blood pressure (SBP).

```{r}
#| label: pca-prep
vars <- c("Age", "BMI", "Waist", "Cholesterol", "SBP")
X <- nhanes[, vars] |> na.omit() |> scale()
pca <- prcomp(X, center = FALSE, scale. = FALSE)  # already scaled
summary(pca)
```

```{r}
#| label: scree-plot
#| fig-width: 7
#| fig-height: 4
var_exp <- pca$sdev^2 / sum(pca$sdev^2)
cum_var <- cumsum(var_exp)

df_scree <- data.frame(
  PC  = factor(paste0("PC", 1:5), levels = paste0("PC", 1:5)),
  var = var_exp,
  cum = cum_var
)

ggplot(df_scree, aes(x = PC, y = var)) +
  geom_col(fill = "#3498db", alpha = 0.8) +
  geom_line(aes(y = cum, group = 1), colour = "#e74c3c", linewidth = 1) +
  geom_point(aes(y = cum), colour = "#e74c3c", size = 3) +
  geom_hline(yintercept = 0.8, linetype = "dashed", colour = "grey50") +
  scale_y_continuous(labels = scales::percent_format()) +
  labs(title = "Scree plot — NHANES health variables",
       subtitle = "Bars: variance per PC  |  Red line: cumulative",
       x = NULL, y = "Variance explained") +
  theme_minimal()
```

```{r}
#| label: loadings
round(pca$rotation, 3)
```

```{r}
#| label: biplot
#| fig-width: 7
#| fig-height: 6
biplot(pca, scale = 0, cex = c(0.4, 0.9),
       col  = c("grey70", "#2c3e50"),
       main = "Biplot — PC1 vs PC2")
```

**Interpretation:** PC1 captures the shared "adiposity + cardiovascular burden" axis (BMI, Waist, SBP load positively). PC2 separates Age from the metabolic variables. This mirrors what FA would find, but the components are labelled by variance, not by psychological constructs.

---

# Guided Practice

::: {.webr-loading-note}
**R is loading in the background** — the exercises below will be interactive once the status bar disappears.
:::

```{webr-r}
#| label: guided-pca
vars <- c("Age", "BMI", "Waist", "Cholesterol", "SBP")
X <- na.omit(nhanes[, vars])
pca <- prcomp(X, scale. = TRUE)

# TASK 1: How much variance does PC1 explain? And PC1+PC2 combined?
var_exp <- pca$sdev^2 / sum(pca$sdev^2)
cat("PC1:", round(var_exp[1]*100, 1), "%\n")
cat("PC1+PC2:", round(sum(var_exp[1:2])*100, 1), "%\n")

# TASK 2: Which variable loads most strongly on PC1?
cat("\nLoadings on PC1:\n")
print(round(pca$rotation[, 1], 3))
```

```{webr-r}
#| label: guided-scree
vars <- c("Age", "BMI", "Waist", "Cholesterol", "SBP")
X <- na.omit(nhanes[, vars])
pca <- prcomp(X, scale. = TRUE)

# TASK 3: Plot the scree plot — how many PCs would you keep?
plot(pca, type = "l", main = "Scree plot")

# TASK 4: Use the biplot to identify which variables are most correlated
biplot(pca, scale = 0, cex = c(0.3, 0.9))
```

::: {.callout-tip collapse="true"}
## Hints
- `pca$sdev^2` gives eigenvalues; divide by `sum(...)` for proportions
- `pca$rotation[, 1]` gives loadings on the first PC
- Look for the "elbow" in the scree plot — the point where adding more PCs gains little
- In the biplot, arrows pointing in the same direction = positively correlated variables
:::

---

# Retrieval Challenge

```{webr-r}
#| label: challenge
# Run PCA on a different set of variables — include Sex as a numeric (0/1) variable
# and explore whether it separates on PC1 or PC2

vars2 <- c("BMI", "Waist", "Cholesterol", "SBP", "Age")
dat2 <- na.omit(nhanes[, c(vars2, "Sex")])
pca2 <- prcomp(dat2[, vars2], scale. = TRUE)

# Project onto PC1 and PC2
scores <- as.data.frame(pca2$x[, 1:2])
scores$Sex <- dat2$Sex

# Compare PC1 scores by sex
tapply(scores$PC1, scores$Sex, mean)
```

::: {.callout-tip collapse="true"}
## Model Answer
```r
vars2 <- c("BMI", "Waist", "Cholesterol", "SBP", "Age")
dat2 <- na.omit(nhanes[, c(vars2, "Sex")])
pca2 <- prcomp(dat2[, vars2], scale. = TRUE)
scores <- as.data.frame(pca2$x[, 1:2])
scores$Sex <- dat2$Sex
library(ggplot2)
ggplot(scores, aes(PC1, PC2, colour = Sex)) +
  geom_point(alpha = 0.3, size = 0.8) +
  stat_ellipse() +
  theme_minimal() +
  labs(title = "PC1 vs PC2 by sex")
```
:::

---

# Why-Questions

::: {.callout-caution collapse="true"}
## Q: "Can I use PC scores as predictors in regression instead of the original variables?"

Yes — this is called Principal Component Regression (PCR). It solves multicollinearity (BMI + Waist from the OLS page) because PCs are orthogonal by construction. The trade-off: PCs are harder to interpret than original variables, and there is no guarantee that the PCs that explain the most variance in X are the ones most predictive of Y. LASSO and Ridge often outperform PCR because they let the response guide variable selection.
:::

::: {.callout-caution collapse="true"}
## Q: "How is PCA different from what I did in STA220 with Cronbach's alpha?"

Cronbach's alpha measures internal consistency — whether items within a scale measure the same construct reliably (high inter-item correlation). PCA finds the direction of maximum variance regardless of reliability. On a well-designed scale like HAD, the first PC will align closely with the sum score (which Cronbach's alpha validates). But PCA on items with low inter-item correlation still produces components — they just explain less variance and are harder to interpret as constructs.
:::

::: {.callout-caution collapse="true"}
## Q: "Should I use the correlation matrix or the covariance matrix?"

Use the correlation matrix (i.e., `scale = TRUE`) when variables are on different scales — which is almost always. Using the covariance matrix means variables with larger variances dominate PC1. If you have five health variables measured in different units (years, kg/m², mmol/L, mmHg), the one with the largest raw variance will drive the first component — an artefact of measurement, not biology.
:::

::: {.callout-note}
## Further reading
James et al. (2021). *An Introduction to Statistical Learning*. — Chapter 12.2
:::
