---
title: "U2 — K-Means Clustering"
engine: knitr
---

```{=html}
<script>document.body.classList.add("unsupervised");</script>
```

```{r}
#| context: setup
#| include: false
library(ggplot2)
library(dplyr)
nhanes <- readRDS(here::here("data/nhanes.rds"))
```

```{webr-r}
#| context: setup
#| include: false
library(ggplot2)
library(dplyr)
nhanes <- readRDS("https://johlobs.github.io/STA230-App/data/nhanes.rds")
```

# Concept: Finding Groups Without Labels

## Supervised vs unsupervised — the key distinction

In Blocks 1 and 2 you always had a response variable Y: SBP to predict, disease to classify. Clustering has no Y. You observe $\mathbf{X}$ and ask: **are there natural subgroups in this data?**

In STA220 you asked a related question: *do HAD items group into depression and anxiety?* Factor analysis answered this by finding latent dimensions. Clustering answers a different version: *do patients group into distinct health profiles?* No labels are assumed — the groups emerge from the data geometry.

## The k-means algorithm

Given $n$ observations and a chosen number of clusters $K$, k-means minimises the **total within-cluster sum of squares (WCSS)**:

$$\text{WCSS} = \sum_{k=1}^{K} \sum_{i \in C_k} \|\mathbf{x}_i - \boldsymbol{\mu}_k\|^2$$

where $\boldsymbol{\mu}_k$ is the centroid (mean) of cluster $k$.

**Algorithm (Lloyd's):**

1. Randomly initialise $K$ centroids
2. **Assign** each observation to the nearest centroid
3. **Update** each centroid as the mean of its assigned observations
4. Repeat steps 2–3 until assignments stop changing

The algorithm converges to a **local** minimum — not necessarily the global minimum. Always run with multiple random starts (`nstart` in R) and take the best result.

## Choosing K — the elbow method

Run k-means for $K = 1, 2, \ldots, K_{\max}$ and plot WCSS against $K$. WCSS always decreases as $K$ increases (more clusters = smaller groups = tighter fit). Look for an **elbow** — the point where the rate of decrease slows sharply.

There is no objectively correct $K$. Domain knowledge matters: do you expect 2 patient profiles or 5? The elbow is a heuristic, not a rule.

## Important: scale your variables

K-means uses Euclidean distance. If Age is measured in years (range 20–80) and Cholesterol in mmol/L (range 2–8), Age will dominate purely because of its larger scale. **Always standardise before clustering.**

---

# Quiz

```{=html}
<div class="sta230-quiz" data-question="K-means minimises within-cluster sum of squares. What does this objective encourage?">
  <div class="q-opt" data-correct="false">Clusters that are maximally separated from each other in space</div>
  <div class="q-opt" data-correct="true">Observations within each cluster are close to their centroid — compact and tightly grouped</div>
  <div class="q-opt" data-correct="false">Clusters of approximately equal size across all groups</div>
  <div class="q-opt" data-correct="false">Clusters whose observations follow a multivariate normal distribution</div>
  <div class="q-exp">WCSS penalises observations that are far from their cluster centroid. Minimising WCSS makes clusters internally compact. It does not directly maximise between-cluster distance, though the two are related. Equal-size clusters and normality are not required — but k-means performs poorly when clusters are very unequal in size or non-spherical in shape.</div>
</div>

<div class="sta230-quiz" data-question="You run kmeans(X, centers = 3, nstart = 1) twice and get different cluster assignments. Why?">
  <div class="q-opt" data-correct="false">k-means is inherently stochastic and cluster assignments never converge to any stable configuration</div>
  <div class="q-opt" data-correct="true">Random starting centroids lead to different local minima and different final cluster assignments</div>
  <div class="q-opt" data-correct="false">k-means randomly resamples the input data before each run, causing the cluster geometry to change</div>
  <div class="q-opt" data-correct="false">k-means requires at least K ≥ 5 clusters for the algorithm to converge to a reproducible assignment</div>
  <div class="q-exp">With nstart = 1, k-means starts from one random set of centroids. Different starting points lead to different local minima. Always use nstart ≥ 20 and set.seed() for reproducibility. R's kmeans() returns the best result across all starts (lowest WCSS).</div>
</div>

<div class="sta230-quiz" data-question="You cluster patients on Age, BMI, Waist, and SBP without standardising. BMI ranges from 15–50 and SBP from 90–200. What is the problem?">
  <div class="q-opt" data-correct="false">K-means cannot process four variables simultaneously — it is restricted to working with only two dimensions</div>
  <div class="q-opt" data-correct="false">K-means automatically forms perfectly spherical, equal-volume clusters regardless of variable scale</div>
  <div class="q-opt" data-correct="true">Large-scale SBP dominates Euclidean distance — clusters reflect blood pressure above all other variables</div>
  <div class="q-opt" data-correct="false">The algorithm will cycle indefinitely between inconsistent cluster assignments and never converge</div>
  <div class="q-exp">Euclidean distance treats all dimensions equally in units. SBP (range ~110) contributes 110² ≈ 12000 to squared distance per unit change; BMI (range ~35) contributes 35² ≈ 1225. SBP numerically dominates. Standardise first so each variable has mean 0 and SD 1, giving equal weight to all variables.</div>
</div>
```

---

# Worked Example: Patient Health Profiles in NHANES

We cluster NHANES patients into groups based on their health profile — no outcome variable used.

```{r}
#| label: kmeans-prep
set.seed(42)
vars <- c("Age", "BMI", "Waist", "Cholesterol", "SBP")
dat  <- na.omit(nhanes[, vars])
X    <- scale(dat)

# Run k-means for K = 2 to 8, record WCSS
wcss <- sapply(2:8, function(k) {
  kmeans(X, centers = k, nstart = 25)$tot.withinss
})

data.frame(K = 2:8, WCSS = wcss)
```

```{r}
#| label: elbow-plot
#| fig-width: 7
#| fig-height: 4
ggplot(data.frame(K = 2:8, WCSS = wcss), aes(K, WCSS)) +
  geom_line(colour = "#3498db", linewidth = 1) +
  geom_point(size = 3, colour = "#2c3e50") +
  scale_x_continuous(breaks = 2:8) +
  labs(title = "Elbow plot — NHANES health clustering",
       x = "Number of clusters K", y = "Total within-cluster SS") +
  theme_minimal()
```

```{r}
#| label: kmeans-k3
set.seed(42)
km3 <- kmeans(X, centers = 3, nstart = 25)
dat$cluster <- factor(km3$cluster)

# Cluster means (back on original scale)
aggregate(dat[, vars], by = list(Cluster = dat$cluster), FUN = mean) |>
  mutate(across(where(is.numeric), round, 1))
```

```{r}
#| label: cluster-plot
#| fig-width: 8
#| fig-height: 5
# Visualise clusters on first two PCs
pca <- prcomp(X)
scores <- data.frame(pca$x[, 1:2], cluster = dat$cluster)

ggplot(scores, aes(PC1, PC2, colour = cluster)) +
  geom_point(alpha = 0.4, size = 0.9) +
  stat_ellipse(linewidth = 0.8) +
  scale_colour_manual(values = c("#e74c3c", "#3498db", "#2ecc71")) +
  labs(title = "K-means clusters (K = 3) projected onto PC1–PC2",
       subtitle = "Ellipses = 95% contours per cluster",
       colour = "Cluster") +
  theme_minimal()
```

**Interpretation:** Three clusters emerge — broadly: younger healthy, middle-aged overweight, older hypertensive. These are data-driven profiles, not clinical diagnoses. The PC projection shows how well-separated the clusters are in the low-dimensional space found by PCA.

---

# Guided Practice

::: {.webr-loading-note}
**R is loading in the background** — exercises will be interactive once the status bar disappears.
:::

```{webr-r}
#| label: guided-elbow
vars <- c("Age", "BMI", "Waist", "Cholesterol", "SBP")
dat  <- na.omit(nhanes[, vars])
X    <- scale(dat)

# TASK 1: Run the elbow method for K = 2 to 8
set.seed(42)
wcss <- sapply(2:8, function(k) kmeans(X, centers = k, nstart = 20)$tot.withinss)
plot(2:8, wcss, type = "b", pch = 19,
     xlab = "K", ylab = "WCSS", main = "Elbow plot")

# TASK 2: Where do you see the elbow? What K would you choose?
```

```{webr-r}
#| label: guided-interpret
vars <- c("Age", "BMI", "Waist", "Cholesterol", "SBP")
dat  <- na.omit(nhanes[, vars])
X    <- scale(dat)
set.seed(42)

# TASK 3: Fit K = 4 and inspect the cluster means
km4 <- kmeans(X, centers = 4, nstart = 25)
dat$cluster <- km4$cluster
aggregate(dat[, vars], by = list(Cluster = dat$cluster), FUN = mean)

# TASK 4: How would you label these four clusters clinically?
```

::: {.callout-tip collapse="true"}
## Hints
- `kmeans(X, centers = K, nstart = 25)$tot.withinss` gives WCSS for a given K
- Use `aggregate()` or `group_by() |> summarise()` to get cluster means
- Look at each cluster's mean Age, BMI, and SBP to give it a clinical label
- There is no "correct" label — interpretation requires domain knowledge
:::

---

# Retrieval Challenge

```{webr-r}
#| label: challenge
# Run k-means with K = 3 on only TWO variables: BMI and SBP
# Then visualise the clusters in a scatterplot
# Compare the cluster separation to the 5-variable result

vars_small <- c("BMI", "SBP")
dat2 <- na.omit(nhanes[, vars_small])
X2   <- scale(dat2)
set.seed(42)
km_small <- kmeans(X2, centers = 3, nstart = 25)
dat2$cluster <- factor(km_small$cluster)

# Your plot here:
```

::: {.callout-tip collapse="true"}
## Model Answer
```r
library(ggplot2)
ggplot(dat2, aes(BMI, SBP, colour = cluster)) +
  geom_point(alpha = 0.3, size = 0.7) +
  stat_ellipse() +
  scale_colour_manual(values = c("#e74c3c", "#3498db", "#2ecc71")) +
  labs(title = "K-means (K=3) on BMI and SBP only") +
  theme_minimal()
# Compare: are the 2-variable clusters similar to the 5-variable clusters?
# What information is lost by ignoring Age, Waist, Cholesterol?
```
:::

---

# Why-Questions

::: {.callout-caution collapse="true"}
## Q: "Is cluster 1 really different from cluster 2? Maybe they just look different by chance."

K-means always produces K clusters — even on completely random data. There is no built-in test for whether clusters are "real". One approach: compare the WCSS improvement from K-1 to K against the same step in randomly permuted data (the **gap statistic**). Another: check if the cluster structure replicates in a held-out half of the data. In practice, assess clusters on clinical interpretability, not just WCSS.
:::

::: {.callout-caution collapse="true"}
## Q: "K-means gave my colleague different clusters on the same data. Is one of us wrong?"

Not necessarily. K-means can find different local optima on the same data, especially when clusters are not well-separated. Both solutions may be valid local minima. Solutions: (1) use `set.seed()` for reproducibility, (2) use `nstart = 25` or more, (3) if results are very sensitive to initialisation, consider hierarchical clustering (next page) which is deterministic.
:::

::: {.callout-caution collapse="true"}
## Q: "Can I use K-means to validate the factor structure I found in STA220?"

Not directly — k-means clusters observations (patients), while factor analysis clusters variables (HAD items). But you can use PCA scores as input to k-means: run PCA on the HAD items → take the first two PC scores → run k-means on patients in that 2D space. Patients who score similarly on both the depression and anxiety dimensions will cluster together. This bridges the two methods.
:::

::: {.callout-note}
## Further reading
James et al. (2021). *An Introduction to Statistical Learning*. — Chapter 12.4
:::
