---
title: "U5 — t-SNE: Nonlinear Dimensionality Reduction"
engine: knitr
---

```{=html}
<script>document.body.classList.add("unsupervised");</script>
```

```{r}
#| context: setup
#| include: false
library(ggplot2)
library(dplyr)
library(Rtsne)
nhanes <- readRDS(here::here("data/nhanes.rds"))
```

```{webr-r}
#| context: setup
#| include: false
library(ggplot2)
library(dplyr)
webr::install("Rtsne")
library(Rtsne)
nhanes <- readRDS("https://johlobs.github.io/STA230-App/data/nhanes.rds")
```

# Concept: Seeing Structure That PCA Misses

## PCA is linear — that is a limitation

PCA finds the directions of maximum variance by a linear transformation of the data. If the true structure is curved or folded — as is common in high-dimensional biological or image data — PCA flattens it out and loses the local neighbourhood relationships.

**t-SNE** (t-distributed Stochastic Neighbour Embedding) is a nonlinear method designed to preserve **local structure**: observations that are close in high-dimensional space are placed close together in the 2D map.

## How t-SNE works (intuitively)

1. In the high-dimensional space, compute a **probability** that observation $j$ is a neighbour of $i$: observations close together get high probability, distant ones near zero.
2. In the 2D embedding, compute corresponding probabilities.
3. **Minimise the KL divergence** between the two probability distributions by adjusting the 2D positions.

The result is a 2D map that reflects the local neighbourhood structure. Unlike PCA, the axes of a t-SNE plot have no interpretation — they are not "components" explaining variance. The distances between clusters are also not directly interpretable.

## Perplexity

Perplexity is the main hyperparameter. It controls how many neighbours each observation "cares about" when building the high-dimensional probability distribution.

| Perplexity | Effect |
|------------|--------|
| Low (e.g. 5) | Focuses on very local structure — can fragment real clusters |
| Medium (e.g. 30) | Standard default — balances local and semi-global structure |
| High (e.g. 100) | Considers more neighbours — blurs local detail, shows broader structure |

**Always run t-SNE with multiple perplexity values** and check whether the structure is stable.

## Important limitations

- t-SNE is **stochastic** — different runs give different layouts. Use `set.seed()`.
- **Cluster sizes and distances in the 2D plot are not meaningful** — a compact cluster is not necessarily more homogeneous than a spread-out one.
- t-SNE is designed for **high-dimensional data** (hundreds or thousands of variables). On five NHANES variables it is demonstrative only; PCA is usually sufficient at low dimensions.
- t-SNE does not produce a transformation that can be applied to new data (unlike PCA).

---

# Quiz

```{=html}
<div class="sta230-quiz" data-question="PCA and t-SNE both produce 2D visualisations. What is the key difference?">
  <div class="q-opt" data-correct="false">t-SNE maximises explained variance; PCA maximises neighbourhood preservation</div>
  <div class="q-opt" data-correct="true">PCA finds global linear directions of maximum variance; t-SNE preserves local neighbourhoods via a nonlinear probabilistic embedding
  <div class="q-opt" data-correct="false">They are equivalent methods — t-SNE is just a faster implementation of PCA</div>
  <div class="q-opt" data-correct="false">PCA works on any data; t-SNE only works on images</div>
  <div class="q-exp">PCA rotates the data so the first two axes capture the most variance globally. Two observations that are close in PC1–PC2 space may actually be separated on a higher PC. t-SNE, by contrast, directly optimises 2D positions to reflect proximity in the full high-dimensional space. t-SNE often reveals cluster structure invisible to PCA, but the axes are meaningless and cluster distances cannot be compared.</div>
</div>

<div class="sta230-quiz" data-question="You run t-SNE with perplexity = 5 and see 20 tiny disconnected islands. With perplexity = 50 you see 3 broad clusters. Which result should you trust?">
  <div class="q-opt" data-correct="false">Perplexity = 5, because more detail is always better</div>
  <div class="q-opt" data-correct="false">Perplexity = 50, because larger perplexity is always better</div>
  <div class="q-opt" data-correct="true">Neither — only structure stable across several perplexity values should be trusted</div>
  <div class="q-opt" data-correct="false">The average visual impression from combining the two plots</div>
  <div class="q-exp">Very low perplexity makes t-SNE focus on tiny local neighbourhoods and can fragment real clusters into many small islands. Very high perplexity approaches PCA-like global structure and may merge distinct groups. Reliable structure appears consistently across a range of perplexity values (e.g. 15, 30, 100). If a cluster only appears at one perplexity, treat it with scepticism.</div>
</div>

<div class="sta230-quiz" data-question="In a t-SNE plot, cluster A is compact and cluster B is spread out. Does this mean cluster A is more homogeneous?">
  <div class="q-opt" data-correct="false">Yes — compact clusters in t-SNE always indicate high internal similarity</div>
  <div class="q-opt" data-correct="true">No — the size and shape of clusters in t-SNE plots are not directly interpretable</div>
  <div class="q-opt" data-correct="false">Yes, but only if perplexity > 30</div>
  <div class="q-opt" data-correct="false">Yes — this is the main advantage of t-SNE over PCA</div>
  <div class="q-exp">t-SNE distorts distances to pack the high-dimensional structure into 2D. The algorithm can make a dense cluster look spread out and a loose cloud look compact, depending on the local neighbourhood sizes. Never interpret the physical size of clusters in a t-SNE plot as a measure of homogeneity. Use silhouette widths or within-cluster distances in the original space for that.</div>
</div>
```

---

# Worked Example: t-SNE on NHANES Health Variables

We use five health variables — note that t-SNE is most powerful for data with hundreds of variables. This example is illustrative; for scRNA-seq or image data the advantage over PCA is much larger.

```{r}
#| label: tsne-prep
vars <- c("Age", "BMI", "Waist", "Cholesterol", "SBP")
dat  <- na.omit(nhanes[, vars])
X    <- scale(dat)
set.seed(42)
```

```{r}
#| label: tsne-run
#| fig-width: 8
#| fig-height: 5
# Run t-SNE with three perplexity values
tsne_15  <- Rtsne(X, dims = 2, perplexity = 15,  check_duplicates = FALSE)
tsne_30  <- Rtsne(X, dims = 2, perplexity = 30,  check_duplicates = FALSE)
tsne_100 <- Rtsne(X, dims = 2, perplexity = 100, check_duplicates = FALSE)

make_df <- function(res, perp) {
  data.frame(Dim1 = res$Y[, 1], Dim2 = res$Y[, 2],
             Perplexity = paste("Perplexity =", perp))
}

tsne_all <- rbind(make_df(tsne_15, 15),
                  make_df(tsne_30, 30),
                  make_df(tsne_100, 100))

ggplot(tsne_all, aes(Dim1, Dim2)) +
  geom_point(size = 0.8, alpha = 0.5, colour = "#3498db") +
  facet_wrap(~ Perplexity, scales = "free") +
  labs(title = "t-SNE on NHANES health variables — three perplexity values",
       subtitle = "Axes have no interpretation — only relative proximity matters") +
  theme_minimal()
```

```{r}
#| label: tsne-vs-pca
#| fig-width: 10
#| fig-height: 4
# Compare PCA and t-SNE (perplexity = 30) side by side
pca <- prcomp(X)

df_pca <- data.frame(Dim1 = pca$x[, 1], Dim2 = pca$x[, 2], Method = "PCA")
df_tsne <- data.frame(Dim1 = tsne_30$Y[, 1], Dim2 = tsne_30$Y[, 2], Method = "t-SNE (perplexity = 30)")

df_both <- rbind(df_pca, df_tsne)
df_both$Age <- rep(dat$Age, 2)

ggplot(df_both, aes(Dim1, Dim2, colour = Age)) +
  geom_point(size = 0.8, alpha = 0.6) +
  scale_colour_viridis_c(option = "plasma") +
  facet_wrap(~ Method, scales = "free") +
  labs(title = "PCA vs t-SNE — coloured by Age",
       subtitle = "t-SNE separates the age gradient more clearly in local neighbourhoods") +
  theme_minimal()
```

**Interpretation:** With only five variables, PCA and t-SNE reveal similar structure. The advantage of t-SNE becomes dramatic with hundreds or thousands of variables — as in the scRNA-seq PBMC3k data in Project 3, where t-SNE separates cell types that overlap completely in PCA space.

---

# Guided Practice

::: {.webr-loading-note}
**R is loading in the background** — exercises will be interactive once the status bar disappears.
:::

```{webr-r}
#| label: guided-tsne
vars <- c("Age", "BMI", "Waist", "Cholesterol", "SBP")
dat  <- na.omit(nhanes[, vars])
X    <- scale(dat)
set.seed(42)

# TASK 1: Run t-SNE with perplexity = 30
tsne30 <- Rtsne(X, dims = 2, perplexity = 30, check_duplicates = FALSE)

# TASK 2: Plot the result
df <- data.frame(Dim1 = tsne30$Y[, 1], Dim2 = tsne30$Y[, 2])
ggplot(df, aes(Dim1, Dim2)) +
  geom_point(size = 1, alpha = 0.5, colour = "#3498db") +
  labs(title = "t-SNE (perplexity = 30) on NHANES") +
  theme_minimal()
```

```{webr-r}
#| label: guided-colour
vars <- c("Age", "BMI", "Waist", "Cholesterol", "SBP")
dat  <- na.omit(nhanes[, vars])
X    <- scale(dat)
set.seed(42)
tsne30 <- Rtsne(X, dims = 2, perplexity = 30, check_duplicates = FALSE)

# TASK 3: Colour the t-SNE plot by BMI — does a gradient appear?
df <- data.frame(Dim1 = tsne30$Y[, 1], Dim2 = tsne30$Y[, 2], BMI = dat$BMI)

ggplot(df, aes(Dim1, Dim2, colour = BMI)) +
  geom_point(size = 1, alpha = 0.6) +
  scale_colour_viridis_c() +
  labs(title = "t-SNE coloured by BMI") +
  theme_minimal()

# TASK 4: Try colouring by Age instead of BMI — what do you see?
```

::: {.callout-tip collapse="true"}
## Hints
- `Rtsne(X, dims = 2, perplexity = 30, check_duplicates = FALSE)$Y` gives the 2D coordinates
- Use `set.seed()` before every t-SNE run for reproducibility
- Colour by original variables to interpret what drives the t-SNE structure
- If the plot looks completely random, try a different perplexity
:::

---

# Retrieval Challenge

```{webr-r}
#| label: challenge
# Run k-means (K = 3) on the t-SNE coordinates
# Then colour the t-SNE plot by cluster assignment
# Compare to k-means run directly on the original 5 variables
# Are the cluster assignments similar?

vars <- c("Age", "BMI", "Waist", "Cholesterol", "SBP")
dat  <- na.omit(nhanes[, vars])
X    <- scale(dat)
set.seed(42)
tsne30 <- Rtsne(X, dims = 2, perplexity = 30, check_duplicates = FALSE)

# Your code here:
```

::: {.callout-tip collapse="true"}
## Model Answer
```r
# K-means on t-SNE coordinates
set.seed(42)
km_tsne <- kmeans(tsne30$Y, centers = 3, nstart = 25)

# K-means on original scaled data
km_orig <- kmeans(X, centers = 3, nstart = 25)

df <- data.frame(
  Dim1 = tsne30$Y[, 1], Dim2 = tsne30$Y[, 2],
  tsne_cluster = factor(km_tsne$cluster),
  orig_cluster = factor(km_orig$cluster)
)

library(tidyr)
df_long <- pivot_longer(df, cols = c(tsne_cluster, orig_cluster),
                        names_to = "method", values_to = "cluster")

ggplot(df_long, aes(Dim1, Dim2, colour = cluster)) +
  geom_point(size = 0.8, alpha = 0.5) +
  facet_wrap(~ method) +
  theme_minimal()

# Cross-tabulation
table(tsne = km_tsne$cluster, original = km_orig$cluster)
```
Clustering on t-SNE coordinates and on original variables often gives different results because t-SNE emphasises local structure. Clustering on t-SNE is reasonable for exploration but the clusters are defined by the t-SNE geometry, not the original variable space.
:::

---

# Why-Questions

::: {.callout-caution collapse="true"}
## Q: "Can I use t-SNE instead of PCA before clustering?"

You can, but be cautious. Clustering on t-SNE coordinates is exploratory and the clusters are defined by the 2D projection, not the original space. Distances in t-SNE are distorted, so k-means on t-SNE will find clusters based on visual proximity in the plot. The standard pipeline for high-dimensional data (like scRNA-seq) is: PCA first to reduce to ~50 components → t-SNE/UMAP on the PC scores for visualisation → cluster in PCA space, not t-SNE space.
:::

::: {.callout-caution collapse="true"}
## Q: "My t-SNE plot looks different every time I run it. Is something wrong?"

No — t-SNE is stochastic. The random initialisation means different runs produce different layouts, though the overall cluster structure should be similar if the perplexity is appropriate. Always use `set.seed()` for reproducibility. If the gross structure (which groups are separated vs merged) changes between runs, the clusters are not stable and you should increase `nstart` or try different perplexity values.
:::

::: {.callout-caution collapse="true"}
## Q: "How is t-SNE different from UMAP?"

Both are nonlinear dimensionality reduction methods that preserve local structure. t-SNE minimises KL divergence between probability distributions in high-D and 2D. UMAP uses a different mathematical framework (topological data analysis) that better preserves global structure — distant clusters in the high-D space remain distant in the UMAP plot, which is not guaranteed with t-SNE. UMAP is also faster and produces deterministic results for fixed seeds. In practice, both are used for visualisation; UMAP is now more common in genomics. See the next page for details.
:::

::: {.callout-note}
## Further reading
van der Maaten & Hinton (2008). *Visualizing Data using t-SNE*. JMLR 9, 2579–2605.
:::
